Parahippocampal neurons encode task-relevant information for goal-directed navigation

Version 1

Abstract

Abstract

A behavioral strategy crucial to survival is directed navigation to a goal, such as a food or home location. One potential neural substrate for supporting goal-directed navigation is the parahippocampus, which contains neurons that represent an animal’s position, orientation, and movement through the world, and that change their firing activity to encode behaviorally relevant variables such as reward. However, little prior work on the parahippocampus has considered how neurons encode variables during goal-directed navigation in environments that dynamically change. Here, we recorded single units from rat parahippocampal cortex while subjects performed a goal-directed task. The maze dynamically changed goal-locations via a visual cue on a trial-to-trial basis, requiring subjects to use cue-location associations to receive reward. We observed a mismatch-like signal, with elevated neural activity on incorrect trials, leading to rate-remapping. The strength of this remapping correlated with task performance. Recordings during open-field foraging allowed us to functionally define navigational coding for a subset of the neurons recorded in the maze. This approach revealed that head-direction coding units remapped more than other functional-defined units. Taken together, this work thus raises the possibility that during goal-directed navigation, parahippocampal neurons encode error information reflective of an animal’s behavioral performance.

Introduction

Navigation to a goal location, such as a food source or home, is a behavior central to the survival of many species. This behavior begins with a serial transformation of sensory inputs and ends in a sequence of motor actions reflective of a movement decision. Between sensation and motor action, several parahippocampal regions represent variables theorized to support the computations needed for goal-directed navigation. These regions include the medial entorhinal cortex (MEC) and pre- and para-subiculum (PrS, PaS), which contain functionally defined neurons that encode an animal’s spatial position, orientation and movement through the external environment [1–6]. Additionally, the PrS and PaS project to the MEC, which is reciprocally connected to the hippocampus [7], a region strongly implicated in supporting memory and navigation [8–10]. However, much of the prior work characterizing functionally defined neurons in MEC, PrS and PaS has considered neural activity recorded as rodents forage for random food rewards in open fields environments. One limitation inherent to such open field foraging behavioral data sets is the lack of pre-defined decision points, which creates challenges to understanding the relationship between neural activity and goal-directed navigation. While several studies have attempted to addresses these challenges by utilizing linear environments, these works typically employed alternation tasks that lack experimental control over decision making junctures [11, 12]. Thus, how parahippocampal neurons encode navigational variables during goal-directed navigation in complex environments remains incompletely characterized.

Goal directed navigation encompasses both spatial and non-spatial coding elements, as an animal must estimate its current body position, head orientation, and running speed and integrate knowledge of behav-iorally relevant goals or sensory cues. Both the MEC and PaS contain large populations of neurons that encode navigationally relevant variables for computing an animals spatial position, including: body position, head direction and running speed. Individual MEC neurons often exhibit a mixture in their selectivity for encoding these variables and, while 50 % of MEC neurons can be classified using tuning curve based metrics, a larger proportion of neurons can be classified using more unbiased statistical approaches [13–15]. In the PaS, 50 % of neurons code for heading direction [6, 16], with coding for position or running speed present in about 40% neurons [15, 16]. MEC has also been shown to contain neurons that encode variables related to non-spatial features of a task or behavior. For example, subsets of MEC neurons encode time-elapsed during immobility [17], change their firing rate or the spatial location at which they are maximally active in response to reward locations [18, 19], and encode an animal’s position in task-relevant space during a non-spatial auditory frequency task [20].

In addition to encoding navigationally relevant variables, parahippocampal regions have been proposed to indirectly support memory through the reciprocal projections to the hippocampal formation [9]. Depending on the task, goal-directed navigational behaviors might require knowledge of remembered cue-reward associations or memory of previous rewarded or behaviorally significant spatial locations [21–23]. At the level of neural activity, these types of memory associations may be facilitated by changes in neural activity that occur in response to changes in environmental sensory cues, such as new visual landmarks, or the presence of a reward [24, 25]. For example, in the hippocampus, place cells that are active in one or a few spatial locations will move the spatial position at which they are maximally active across different open-field environments, a phenomenon referred to as ‘global remapping’. Place cells can also change their firing rate, a phenomenon referred to as ‘rate remapping’, which can occur with global remapping or independent of global remapping when changes to the environment are small [24, 26–28]. In the hippocampus, both global and rate remapping have been observed in response to changes in environmental sensory cues, such as the color or shape of the spatial environment, and the presence of reward [29–31]. Remapping events observed in the hippocampus have been shown to occur in concert with changes MEC spatial firing patterns, at both the correlational level (place cell remapping correlates with re-orientation of MEC grid-cells) [32], and at the causal level (inactivation of MEC coincides with hippocampal global remapping) [33]. Both global and rate remapping measurements provide, at the population level, a quantification of stability in the neural representation of an environment through spatial correlations [24]. However, remapping might be indicative of flexibility in the representation of space, other behaviorally relevant variables, or experience [28]. Thus, quantification of remapping can be experimentally leveraged to examine whether spatial representations flexibly adapt to behaviorally relevant factors such as distinct cue-reward associations that guide decisions made during goal-directed navigation.

Neurons in MEC also exhibit global and rate remapping between different spatial environments. Moreover, MEC neurons also remap in response to variables that flexibly change during goal-directed navigation, including: reward locations, running speed and an animal’s trajectory. The presence or change in the position of an unmarked remembered reward location can evoke both rate and global remapping in position and orientation encoding cells in MEC [18, 19]. Changes in an animal’s running speed have been observed to correlate with both rate remapping [34, 35] and global remapping in MEC [36]. In structured linear track mazes, in which animals perform spatial alternation tasks, MEC exhibits trajectory dependent remapping for the same spatial location [11, 12, 37]. However, while previously observed trajectory dependent remapping is indicative of prospective coding during stereotyped behavior (e.g. alternation), it remains unknown how MEC, PaS and PrS spatial patterns change in tasks with no trial-to-trial predictability or experimenter controlled cues and goals in the same environment. Moreover, while work in the hippocampus has shown that the degree of trajectory dependent remapping correlates with behavioral performance [38], and causal evidence revealed that shifted spatial representations correspondingly shifts the expected location of reward [21], such neural-behavior relationships have not been examined in the upstream MEC, PaS or PrS.

Here, we examined single unit neural activity in MEC, PrS, PaS as rats performed a visually cued navigational task that incorporated both goal-directed navigation and random foraging. We found that a large percentage of neurons changed their firing activity as a function of cue identity, and that the extent of separability (as measured through remapping) in this firing activity correlated with behavioral performance. Furthermore, using encoding models of rate and global remapping, we found that changes in neural activity in the task were sufficiently explained by rate remapping. Interestingly, neural representations were also separable contingent on the presence or absence of reward at the goal well. Finally, single-unit recordings during open field foraging revealed that neurons that encoded head-direction remapped more strongly than other functionally defined units. These results together point to the MEC and upstream PrS/PaS as nodes in the transmission of behaviorally relevant variables during goal-directed navigation.

Results

Tree-Maze behavior and electrophysiological recordings

To examine whether MEC, PrS and PaS neurons encode variables related to goal directed navigation, we designed a linear maze with two decision junctions, referred to here as the Tree-Maze (Fig. 1). Rats (n = 5) learned to navigate a sequence of wells to receive liquid milk rewards (Fig. 1a). There were four possible Goal-wells (G1-G4), and subjects received double rewards for triggering the correct Goal-well. Trials started at the Home-well (H). After animals triggered the Home-well (H), a visual cue turned on that indicated the correct branch to navigate after the Decision-well (D) (RC = Right-Cue, purple, Goal-wells G1-G2; LC = Left-Cue, green, Goal-wells G3-G4) (Fig. 1b). For the correct branch, only one Goal-well contained reward, which was randomly chosen on each trial. Thus, the task combined goal directed navigation (selecting the correct branch) with random foraging (selecting the first Goal-well to visit). After navigating correctly, or incorrectly (no reward and a 10 second timeout), the subject had to navigate back to the Home-well to initiate a new trial. The presence of two possible goal locations after the decisions kept subjects engaged, and dissociated the decision from the spatial location of the reward. For analyses, we segmented the maze into zones (Fig.1a), such that the location of the subject at any given time-point could be localized to a single zone (Fig. 1c).

Goal directed navigation on the Tree-Maze task a. Top view of the Tree-Maze layout and segmentation (indicated by upper and lower case letters) used for analyses (height = 1.4 m, width = 1.2 m). The LED cue panel was located at the end of the maze (flattened for illustration). Colored cue cards on the side of the maze indicate the two cue types by trial (Purple = Right-Cue [RC], Green = Left-Cue [LC]). Possible reward locations denoted by capital letters (Home = H, Decision = D, Goal = G1-G4). Lower case letters correspond to the segments shown on the right side of panel c. Reward wells highlighted in blue. b. Colored lines indicate example trajectories of a rat, 5 trials per panel. L Dec indicates a trajectory towards the left branch and R Dec indicates a trajectory towards the right branch. Left column (Left/Green cue), trajectories to G3 (top) and G4 (bottom) for reward. Both of these were correct navigational decisions and were rewarded. Right column (Right/Purple cue), trajectories to G2 (top) and G3 (bottom), only the top trajectories resulted in reward at a goal. c. Binary trajectory segmentation time window. At any given time-point, the subject can only be at one location in the maze (indicated by the white bins). Bottom axis indicates trial start times (seconds). Top axis indicates the trial number (tr), colored coded by cue. Left/Right axis indicates the identity of the segment, lower case letters on the right correspond to those in panel a. Blue and red dashed vertical lines indicate the end of a trial (blue = correct; red = incorrect). d. Task performance by subject. Each dot corresponds to a session by subject. For , dots indicates the subject mean.

We trained 5 rats on the Tree-Maze task (see Methods-Behavioral Training). Subjects took an average of 25 sessions to reach the performance criterion for surgery (# sessions to criterion, range = 11-50; criterion: accuracy ≥ 70% and # trials ≥ 80; Sup. Fig. 1). Post-electrode implantation (Histology - Sup. Fig. 2, # units by subject Sup. Fig. 3), criteria for inclusion in the analyses were: # units ≥ 1, # trials ≥ 50 (across subject median accuracy = 72.6%, median number of trials per session = 110) (Fig. 1d). The cue on each trial was random, which allowed us to keep track of performance on Switch trials (in which the cue on trial N+1 was different than the cue on trial N, Sup. Fig. 1). Indicating that subjects were adept at the task and not simply following an alternation or exploitative strategy, the mean Switch trial performance was similar to non-Switch trials (Switch trial median accuracy = 76.9%; median number of Switch trials per session = 54). Together, the behavioral requirements of the task and structure of the Tree-Maze allowed us to then examine whether neural activity corresponded to different task dimensions (e.g. cue, decision, reward, spatial position).

Visual cues evoked spatial remapping

To investigate the effect of the cue on the spatial firing profiles of single units (i.e. neurons), we generated spatial maps by cue condition for outbound trajectories (H-well to G-well, Fig. 2a. For each unit, the spatial locations of spikes along trajectories were used to obtain trial summary metrics for each maze segment as a function of cue (Fig. 2b; Sup. Fig. 4-5). Across trials, the difference on cue coding by maze segment was quantified by the Mann-Whitney Z transformed U statistic (UZ), (Fig. 2b,c). In the scenario in which the cue had no effect on spatial coding, the distributions of UZ scores will be close to zero for all maze segments (Fig. 2c). On the other hand, the statistic will be negative if firing rates for the Left-Cue (LC) trials were greater than for Right-Cue (RC) trials, and positive if the firing rates for the Right-Cue (RC) trials were greater than for the Left-Cue (LC) trials. Using a linear mixed effects model we found a main effect of Segment with follow-up pairwise comparisons showing a left>stem>right pattern (LMEM; accounted for within subject variance and repeated measures for each session, Segment Likelihood Ratio Test , p = 3.07e-19). Hence, across the population there was a significant bias for Right-Cue trials to be associated with greater firing rates than Left-Cue trials on the left segment of the maze, and for Left-Cue trials to be associated with greater firing rates than Right-Cue on the right segment of the maze. In both of these instances, the increased firing rate thus follows an incorrect decision, that is, trajectories that did not lead to a reward were the trajectories associated with an increased firing rate.

Spatial remapping was associated with the visual cue and correlated with task performance. a. Example session trajectories for all LC (Left-Cue) and RC (Right-Cue) trials. b. Four example single-units. For each example unit: Top row, left and middle, outbound trajectories separated by cue (red dots = spikes). Top-right, trial median activity by cue and maze segment (left, stem, right), numbers at the top indicate the Mann-Whitney Z transformed U statistic for the difference between the RC and LC trial distributions Uz. Bottom row, mean spatial rate maps by zone and cue, color coded for minimum (firing rate [FR] = 0, blue) and maximum (yellow) values. Bottom right, re-sampled distribution of correlations between RC and LC maps in blue for that unit, and in grey the corresponding null distribution. Remapping score is the mean remapping score for each unit. c. Distributions of UZ scores for all recorded units by maze segment. Purple means higher FR for RC than LC, Green means higher FR for LC than RC. Note the higher FR for RC on the left segment (far left) and higher FR for the LC on the right segment (far right). d. Distribution of mean remapping scores by unit, note the negative shift in the distribution of scores. e. Scatter-plot between the task performance on a given session pse and the cue remapping scores for recorded units in that session. Size and color of dots scale with the x axis for illustration. Regression line in red with a CI95% band. Kendall correlation score between behavior and remapping score shown. f. Scatter-plot between pse and the mean remapping score across units recorded in a given session . Size of dots indicate number of co-recorded units, color codes correspond to different subjects. Regression line and corresponding CI95% band shown in grey. g. Like (f) but with neural population correlation, composed of the spatial rate maps for all recorded units in a session. g. Correlation between pse and by subject, with bootstrapped standard deviation. is the across subject mean.

Next, to quantify the degree of spatial firing rate changes, we generated spatial zone rate maps as a function of cue. Importantly, these maps were generated from balanced re-sampling of correct and incorrect trials. Concretely, the LC maps were generated by sampling an equal number of correct and incorrect trials from all the trials that had a LC (Fig. 2b). A distribution of scores was computed by taking the Kendall correlation, τ, between LC and RC spatial maps for each trial re-sampling instance (Fig. 2b). An appropriate null distribution was generated by sampling equal numbers of LC and RC trials from even and odd trial sets, computing the spatial correlation, and then comparing the resulting null correlation score to the cue correlation scores to generate a remapping score . Across units we observed a significant negative shift in (LMEM: mean=−0.67, p = 0.003, CI95%=(−1.105,−0.225)) (Fig. 2d). This negative shift indicated that the cue induced a significant level of remapping in firing patterns when compared to a null distribution.

Remapping of spatial representations relates to greater task performance

Given that trajectories after incorrect decisions tended to elicit an elevated FR (Fig. 2c), we hypothesized that the cue remapping score would related to a subject’s performance in the task. To examine this, we related the cue remapping score of each unit to the corresponding task performance for the recorded session pse (Fig. 2e). There was a significant negative relationship between performance accuracy on a given session pse and the degree of cue remapping calculated for single units (Kendall’s τ=-0.27; LMEM: , CI95%=(−0.15,−0.09), , p = 8.84e — 13). Given that this effect was observed for the majority of units, we hypothesized that the effect should be reflected in the mean of the cue remapping scores for co-recorded units in a session (Fig. 2f). Indeed, when considering co-recorded units in a session, the correlation between cue remapping and behavioral performance was even more pronounced (Kendall’s τ=-0.41; LMEM: , , p = 1.4e — 4, CI95%=(−0.33,−0.11)). Population level correlations showed the same effect, (Fig. 2g). At the individual subject level, this correlation between performance accuracy and cue remapping was observed in all but one subject, which had a small number of units (Fig. 2h). These same effects were observed on multiple versions of these analyses (Sup. Fig. 8,9). Together, these results demonstrate cue-based remapping of parahippocampal spatial firing patterns in a goal directed navigation task, with the extent of remapping significantly correlating with behavioral accuracy in the task.

Next, we sought to understand the extent and the manner in which the visual cue was encoded in the FR of individual units during the task. We compared three different linear encoding models: (1) a zone/position model Z, (2) a cue rate-remapping model Z + C, and (3) a cue global-remapping model Z × C. The zone only model Z, simply predicted a neuron’s FR based on the zone/position the subject was occupying at a given time. The cue rate-remapping model Z + C, also received a cue identity input, acting as gain on the model to uniformly modify the predicted FR across all zones as a function of cue (Fig. 3a,b). The cue global-remapping model Z × C doubled the number of input zones, such that only one set of zones could be active on a given cue trial. This approach thus modeled the possibility of a complete re-shuffle of the spatial firing patterns as a function of cue (Sup. Fig. 12). For each unit, the cross-validated coefficient of determination R2 was used to asses model performance (individual scores are shown in Fig. 3c). Note that a negative R2 indicates that the mean firing rate of the neuron explained more activity variance than the fitted model (on held-out test data). A subtle but consistent difference in model scores was apparent between the rate- and global remapping model (Fig. 3c,d), such that the rate remapping model outperformed the global remapping and position only model (LMEM: , p = 1.24e-209) (Fig. 3d). The overall single unit scores showed a mean R2 of 12% for the Z + C model.

Cue modeling revealed rate remapping and trial-wise correlations to behavior. a. Linear zone encoding model with cue Z + C, at a given sample time t, the current position of the animal and the cue identity is multiplied by learned weights to predict FR for each recorded unit . b. Example time window of Z[t], the true FR in black V[t] and the predicted FR in red . c. Model comparison between three types of zone encoding: Z → only zones, Z + C → zones + cue, Z × C → a set of zones for each cue. Each dot is a unit, blue dots were negative R2, red scales with R2 and y-axis. d. Model comparison scores. Y-axis is the Mann-Whitney Z transformed statistic for comparing the R2 on test folds. Colorbar indicates the median difference in R2 across test folds. e. Performance of linear zone decoder models V (circles) and V + C (squares). Y-axis is the error distance in cm between the predicted and true zone, X-axis is the linearized Tree-Maze zones displayed as H (Home-well) to D (Decision-well) to i (second intersection/branching) to G (Goal-well). Linearization achieved through averaging the equivalent trajectories towards the goal. The hue shade provides groupings of sessions according to number of co-recorded units (both isolated an MUA included in these analyses). f. Performance of linear decision decoder by zone, with color indicating number of units. Note the sharp decision well split in the performance. BAC=balanced accuracy. e. Correlation between the subject’s performance and the model by zone. Color groupings as in (f.). Model performance is the comparison between the output of the decision decoder and the true identity of the cue, the same computation used to assess a subject’s performance.

To further understand how the encoding of the cue across units could modulate the strong position coding present in parahippocampal neurons, we trained a linear decoder to predict the position of the subject using the neural population activity patterns with and without the identity of the cue (V + C and V models, respectively, Fig. 3e). On held-out test data, we observed significant modulation with the cue as an input, reflected by decreased errors (the distance between the predicted and true position), but only in sessions with high number of units (LMEM: interaction between model and units groupings: , p = 1.9e-4). Additionally, the addition of cue information was most beneficial in reducing errors post-decision (LMEM: triple interaction of model, unit grouping and pre-/post-decision marker: , p = 2.9e-47). Next, we trained a classifier to predict the Left/Right decision of the subject using the population activity (Fig. 3f). Prospective decoding of the decision prior to the decision point (stem) was at chance, while it steadily increased beyond the decision point across sessions, more so in sessions with more than 8 units (LMEM: interaction of unit groupings and pre-/post decision marker, , p = 3.28e-78). The decoding of decision is reflective of the strong position coding in parahippocampal units, and also indicative of the behaviorally relevant information content of these neurons. An important note is that the number of units in a session heavily influenced the ability of the decoder to predict an upcoming Left/Right decision in the stem. Concretely, 19/42 sessions had above chance decoding with 8 or fewer corecorded units, while 23/36 sessions with 20 or more co-recorded units had above chance decision decoding. Finally, because the decoder has trial-wise predictions of an animal’s decision Left/Right, we tested the extent to which parahippocampal representations matched the subject’s behavior (Fig. 3f). We compared decoder outputs (Left/Right decisions) to the identity of the cue, thus providing a measure comparable to the performance of subjects on the task. With this approach, we relate the performance of the subject to the model’s performance with a LMEM, with unit groupings and pre-/post-decision marker as covariates, and found a triple interaction between these predictors (, p = 3.61-10). The interaction can be broken down as follows: post-decision decoder predictions related to the subject’s performance and was modulated by the number of co-recorded neurons. These analyses demonstrate that the parahippocampal representations are not only reflective of behavior on the mean sense, as assessed through spatial remapping, but also at the individual trial-level.

Absence of reward leads to higher activity and spatial remapping

After the subjects navigated to Goal-wells, they received reward contingent on whether they made the correct navigation decision. Inbound trajectories from the Goal-wells to the Home-well could thus be separated by rewarded (correct trials) versus unrewarded (incorrect trials) (Fig. 44a). Based on previous data [18, 19], we posited that reward could also elicit firing rate changes in the spatial representations of parahippocampal neurons. We quantified firing rate changes for each unit as a function of reward by maze segment (Fig. 44b,c). Across units, we found that unrewarded trials tended to have elevated firing rates across all maze segments (proportions: left=386/565, stem=340/565, right=412/565; LMEM: segment ,p = 7.9le-12).

Given the firing rate differences observed between rewarded and unrewarded trials, we expected to see differences in position coding for the spatial maps corresponding to these trials sets. For each unit we generated a distribution of correlations between rewarded and unrewarded inbound spatial trajectories (bootstrapped trials with equal numbers of LC and RC). The null distribution was then the correlations between even and odd inbound trial trajectories (bootstraps having an equal number of rewarded and unrewarded trials). As with the cue, we quantified the difference between these distributions for each unit to obtain a remapping score ZΔτ, Fig. 4b,d. Across units, we observed a significant negative shift of the remapping score, which indicates that spatial maps changed based on the receipt or absence of reward (LMEM: mean=−1.03, CI95%=(−1.6,−0.46), p = 4.1e-4) (Fig. 4d).

Absence of reward leads to higher activity and spatial remapping. a. Example session RW (reward) and NRW (no-reward) trials. b. Four example units (i-iv). Top rows, inbound trials trajectories by RW/NRW (red dots = spikes). Top-right, trial median activity by RW and segment in the maze. Firing rate difference score Uz as described in the main text. Bottom rows, mean spatial rate maps by zone and RW, minimum FR=0. Bottom right, re-sampled distribution of correlations between RW and NRW maps in blue for that unit, and in grey the corresponding null distribution. Remapping score is the mean remapping score for each unit. b.i Session is the same as in panel a., other units from different sessions. c. Distributions of Uz scores for all recorded units by maze segment. Blues means higher FR for RW than NRW trials, Red higher NRN than RW. d. Distribution of mean remapping scores by unit, note the negative shift in the distribution of scores. e. Scatter-plot between a session’s task performance pse and the remapping scores for recorded units. Size and color of dots scale with the x axis for illustration. Regression line in red with a CI95% band. Kendall correlation score between behavior and remapping score shown. f. Scatter-plot between pse and the mean remapping score across a sessions units . Size of dots indicate number of co-recorded units, color codes correspond to different subjects. Regression line and corresponding CI95% band shown in grey. g. Like (f) but with neural population correlation, composed of the spatial rate maps for all recorded units in a session. h. Correlation between pse and by subject, with bootstrapped standard deviation. is the across subject mean.

Furthermore, because rewarded trials are correct trials, we hypothesized that as with the cue, remapping scores should relate to task performance. The extent of reward related remapping correlated with performance accuracy for a session at the level of individual units, across co-recorded units, and individual subjects (Fig. 4e–h, Sup. Fig. 10,11). Statistical analyses of these results were indicative of a strong relationship between reward remapping and task performance: unit-level (Kendall’s τ=-0.28; LMEM: , CI95% = (−0.18,−0.11), , p = 3.06e-20); across co-recorded units (τ=−0.43; LMEM: slope=−0.28, CI95% = (−0.38,−0.19), , p = 3.28e-8). Similar to cue induced remapping, the extent to which units remapped in response to reward receipt significantly correlated with behavioral accuracy in the task.

As with the cue, we modeled the firing rate activity of individual neurons through three linear encoding models: (1) a zone/position model Zi, (2) a reward rate-remapping model Zi + R, and (3) a reward global-remapping model Zi × R. Whsere the i subscript indicates inbound trajectories, Fig. 5a,b. Similar to the cue models, we found that the activity of most units were better fitted by a rate-remapping model (Fig. 5c,d,e) (LMEM: UΔR2 Model , p = 1.98e-146) (Fig. 5d). Thus these set of results indicate that parahippocampal neurons (as a population) exhibit increased firing rates during inbound trajectories that follow unrewarded trials.

Encoding model of reward remapping. a. Linear zone encoding model with reward Zi + R, at a given sample time t, the current position of the animal and the reward identity is multiplied by learned weights to predict FR for each recorded unit . b. Example time window of Z[t], the true FR in black V[t] and the predicted FR in red . c. Model comparison between three types of zone Encoding during inbound trajectories: Zi, Zi + R, Zi × R. Each dot is a unit, blue dots were negative R2, red scales with R2 and y-axis. d. Model comparison scores. Y-axis is the Mann-Whitney Z transformed statistic for comparing the R2 on test folds. Color-bar indicates the median difference in R2 across test folds.

Given the observations of spatial remapping for both cue and RW condition contingencies, we sought to understand the relationship between these observations. First, we found that remapping scores were correlated (Sup. Fig. 13) across individual units and at the population level. Second, because correct/incorrect trial contingencies largely contributed to how cue and RW spatial maps were constructed, we partitioned the data into correct/incorrect coding and sought to find a the relationship across cue and RW, (Sup. Fig. 14). With this approach, found that incorrect coding units were more likely to overlap between cue and RW coding units than correct. However, remapping scores didn’t differed between these groupings. Third, we conditioned the activity of correct and incorrect inbound trajectories based on the activity of correct and incorrect outbound trajectories (Sup. Fig. 15). This analyses was motivated based on the structure of the trials: rewarded (correct) inbound trajectories followed correct outbound trajectories and that unrewarded (incorrect) inbound trajectories followed incorrect outbound trajectories. We found that incorrect>correct activity levels on outbound trajectories predicted incorrect>correct activity levels on inbound trajectories, suggesting that error signaling activity persisted through the inbound phase of the trial. Together, the remapping observations provide a quantification of this error or mismatch-like signal through different phases of the task and in the context of spatial representations.

Units encoding head direction during open field foraging are more likely to remap

In addition to the Tree-Maze, single units were recorded from subjects while they foraged for randomly scattered food rewards in a rectangular open arena (open-field foraging, OF, Fig.6a). Removable panels forming the floor of the arena were placed on top of the Tree-Maze, thus keeping peripheral visual cues relatively constant between the two environments, with the main difference coming from the 12 cm in gained height for the OF environment. A model-based encoder approach was used to quantify the extent of navigational coding for each unit along the behavioral measures of speed, head-direction or position (Fig.6). The firing-rate predictions of these individual models was used to predict the firing rate of single units on test data in an aggregate model (Fig.6a,b). Across units, the aggregate model produced better results than individual models (LMEM: main effect of model , p = 1.33e-122, , p = 3.44-190) (Fig. 6c,d). Units across the Tree-Maze and OF tasks were matched through a waveform matching algorithm Fig. 7a. In total 217 units were matched across 76 OF and 82 Tree-Maze sessions Fig. 7b. The set of model coefficients (speed, head-direction, position) for each matched unit were clustered into three groups (Fig. 7c–e). Notably, the obtained unit cluster identified a main dimension of separation between clusters in the form of a coding trade-off between head-direction and speed (Fig. 7d,e). That is, units that coded strongly for head-direction (cluster 0, blue, n=98) coded poorly for speed, while units that coded strongly for speed coded poorly for head-direction (cluster 2, green, n=59). For completeness, OF tuning score metrics were also computed for these clusters with results qualitatively matching the model coefficients (Fig. 7f).

Modeling navigational/spatial variables in neural coding during open-field foraging. a. Neural responses of four example units for subjects foraging an open-field (OF) arena [1.3m x 1.5m]. For each unit sub-panel (c0-c3): top-left, firing-rate map (number is the peak FR); top-right, head-direction tuning curve, color indicates FR magnitude by angle; bottom-left, speed tuning curve (s=speed); bottom-right, model-based variance explained on test data (R2) by variable (h=heading-direction, p=position, a=aggregate model). b. Model-based responses by variable for a selected test time-window for units c0-c3. Each row corresponds to a different model prediction , the true firing-rate (fr) for that unit, or at the top the color-coded time-window (t). Top-right, the data on which the model was trained is in grey and super-imposed is the test-window color-coded by time and with firing rate magnitude in increasing dot-size. Other heat-maps are the resulting firing-rate maps generated for the test trajectory. Model predicted rates are shown with colors matching a., with the background dotted line being the true (fr) a.. c. Population level (R2) for train (tr) and test (te) sets. d. Population level firing-rate map Pearson correlation rp between true m and predicted maps . Note that for both metrics, the aggregate model and the position model produced the best results. e. Relationship between coefficients on the aggregate model by unit. The color corresponds to the model’s training set R2.

Matched units across tasks reveal that head-direction coding units remap the strongest. a. Procedure for matching units across tasks (OF-open field, TM-tree maze). Top, 6 tetrode waveforms color-coded by matched units; bottom-left, fitted Gaussians to the dimensionally reduced unit waveforms for that tetrode across matched depth sessions (grey, unmatched units); bottom-right, symmetric confusion error matrix across units, threshold for matching PE <= 0.5. b. Venn diagram of matched units across tasks. c. UMAP clustering of the aggregate model coefficients for matched units. d. Head-direction (h) vs speed (s) coefficients formed a clustering subspace. e. Model coefficients of aggregate model used for finding clusters (p=position). Horizontal lines are the population mean, and error bars are the mean’s 95% CI. Statistics not performed, as the clusters were fitted from these parameters. f. Tuning metric scores (p=split-half rate-map correlation; s=speed-score; h=resultant-vector length score). Paired statistics through Mann-Whitney U tests (*=p < 0.05, **=p < 0.01, ***=p < 0.001, ****=p < 1e-4) h. Remap scores by OF cluster. g. TM model scores by OF cluster.

Next, we sought to relate how these clusters of units defined in the OF foraging task related to the results observed in the Tree-Maze. We observed that on this matched subset of neurons, significant remapping scores were still observed for both cue and RW contingencies, and that units in cluster 0 remapped more than the other clusters (LMEM: cluster x remap interaction , p = 0.032; cluster , p = 0.47), (Fig. 7g). Similarly, Tree-Maze encoder rate/global remapping scores were higher for cluster 0 (LMEM: cluster , p = 9.62e-4) (Fig. 7h). Additionally, we explored the extent to which these clusters related to other aspects of coding in the Tree-Maze like correct/incorrect coding and cue/RW coding by segment (Sup. Fig. 16-17). We found that these functionally defined units had some overlap with different aspect of the maze, suggesting a small degree of heterogeneity in coding. However, most units in the clusters did not overlap with cue/RW or correct/incorrect coding. Taken together, units in cluster 0, that corresponded to stronger head-direction coding, were more prone to remap for both cue and RW than units in cluster 1 or 2. Hence, we leveraged Open-Field recordings to identify head-direction coding units as showing changes of firing patterns in the Tree-Maze, with those changes being reflective of task performance.

Discussion

During Open-Field foraging, individual MEC, PrS and PaS neurons encode an animal’s position, orientation and speed but the firing patterns of neurons in these parahippocampal regions during goal-directed navigation has been less well characterized. Here, we examined the firing patterns of hundreds of parahippocampal neurons as rats navigated a visually-guided goal directed Tree-Maze task. Rats successfully learned to use a visual cue to navigate the left or right branch of a multi-decision point linear maze to receive reward. We found that many neurons exhibited higher firing rates after incorrect left or right branch choices, which led to remapping of spatial firing rate maps. Further, the degree of remapping correlated with behavior, such that higher remapping scores corresponded to better task performance. Using encoder models, we then demonstrated that rate remapping, over global remapping, was sufficient to explain the observed changes in spatial firing rate maps. Finally, by using the navigational coefficients of encoder models fit to Open-Field foraging data, we further revealed that head direction coding cells showed the largest magnitude of remapping on the Tree-Maze. Conjunctively, these results demonstrate that parahippocampal representations of space and navigation, under goal-directed experimental conditions, project a flexible code reflective of behavior.

Neuronal encoder models serve as an important tool for identifying single-unit activity coding during open-field foraging that avoid imposing a prior on the shape of tuning curves or the generation of coding indices (e.g. tuning curve scores) [13]. Moreover, these models allow for the quantification of mixed selectivity in the coding for spatial/navigational variables, providing a more nuanced understanding of the possible variables a neural population encodes. Here, we fit linear encoder models that used speed, heading-direction and position during open-field foraging to predict each unit’s firing rate activity. Consistent with prior work, we find strong position coding, that is often complementary with coding for head-direction or speed ([1, 13, 14, 39]). Although there are differences in the proportion of cells that encode position, head direction and speed between MEC, PrS and PaS (e.g. PrS and PaS have higher proportions of head-direction tuned cells), these regions share a common substrate of spatial/navigational coding (position, head-direction, speed) [6, 15, 16]. We further found a relationship between navigation variables after clustering that revealed an axis of separation along the head-direction and speed coefficients, such that these variables were negatively related. This anti-correlation between speed and head-direction coding replicated recent observations across parahippocampal regions (MEC, PaS and PrS) [15]. Thus, as a function of speed, the head-direction signal monotonically decreases at faster speeds, consistent with the observation that subjects are less likely to turn when they run fast. Thus, one possibility is that the head-direction signal in the parahippocampal region reflects a behavioral state related to navigational choice or the lack of commitment to a particular navigational route.

Prior works point to the MEC as a key node supporting spatial decisions, as it’s inactivation led to performance deficits in visual-scene based spatial decisions [40], and subsets of MEC neurons encode the near future position of the animal (<1 second in the future) [2, 41]. Nonetheless, in terms of navigation behavior, parahippocampal neural activity after incorrect decisions has been understudied. Depending on the task, there are often very few error trials or multiple confounds that prevent substantial quantification of errors (e.g. loss of attention or satiation). Error or mismatch signals conform a fundamental computation, that informs immediate behavior, learning and neural plasticity [42, 43]. Our task structure enabled us to more rigorously consider error trials and the activity of MEC, PaS and PrS neurons before and after such trials. Given that many units exhibited higher firing rates on incorrect trials, leading to rate remapping, we hypothesized that the extent of remapping should correlate with behavioral performance. Indeed, we observed strong correlations between neural remapping and behavioral performance across multiple levels of analyses: individual units, across units, population correlations, individual subject level analyses and a neural population decoder model that mimicked the subject’s behavior (Figs. 2–4), as well as in control analyses that included multi-unit activity or the removal of periods of reward and/or immobility (Sup. Figs. 8-11). The robustness of this observation suggests that cortical regions encode error, expanding the possible brain regions involved in error signalling beyond downstream regions, like the basal ganglia, that prior work has shown carry error signals [42, 43].

Additional support for the idea of a mismatch signal in the parahippocampus comes from prior works that observed changes in the hippocampus related to correct or incorrect decisions. For example, hippocampal place-cells showed more efficient representations of sequences during correct trials than during incorrect [44], and non-place cells show higher rates during correct versus incorrect trials [45]. Moreover, in the T-maze, the extent of discriminability for upcoming left or right trajectories in the firing rate of hippocampal cells active on the stem of the T-maze (i.e. splitter cells) correlated with behavioral performance [38]. Of note however, we did not observe a significant population of ‘splitter cells’ in the central stem of the maze, possibly because the Tree-Maze differs from previous T-or Y-Maze in that there is a reward well at the decision point [37] [11, 12]. Even so, in our task, after training, subjects were more likely to be running on the left branch of the maze when the green visual cue (LC) was present, but after an incorrect decision, they were running on the left branch of the maze when the purple cue (RC) was present. These incorrect spatial decisions led to increase firing rates on the majority of units, and persisted on the inbound trajectories (quantified by rewarded vs unrewarded, Sup. Fig. 15). Notably, units coding for correct decisions on outbound trajectories rarely overlapped with units coding for correct decisions on inbound trajectories, while neurons coding for incorrect decisions often overlapped in their coding for inbound and outbound trajectories (Sup. Fig. 14). These observations are generally consistent with the idea that a mismatch signal could be transmitted by MEC, PaS and PrS neurons via a change in firing rate. However, it remains an open question as to where this type of mismatch signal originates such that it is broadly reflected across parahippocampal regions.

A growing set of studies have started to show that spatial and navigational codes in the parahippocampal region are more flexible than previously thought. Observations of spatial remapping in the MEC have been seen after significant changes to the environment in an open-field arena [18, 46], provoking a coordinated re-organization of grid-cells and seemingly random re-organization of other spatially selective cells. Remapping was also observed to be more likely at low running speeds [36], switching between stable spatial representations. These observations would generally be considered global remapping, as the activity fields in space change position. Other studies, however, have shown changes in firing rate and small shifts in fields near rewarded locations during foraging [18, 19]. Our data, including deep MEC, points to changes in firing rate activity on the same spatial location (rate remapping) being sufficient to explain the differences in firing rate observed between conditions (both for cue and reward). The rate remapping result is again consistent with the idea of a mismatch or error signal acting as a gain on individual neurons. Our results demonstrate that through the use of a dynamic environment we increased the behavioral demands on the animal, providing a behavioral assay to contrast parahippocampal neural activity. Further, by recording from the well characterized parahippocampal navigational circuit, we were able show that there are additional dimensions of representation or computations only elucidated through the use complex behaviors.

Supporting information

eLife Assessment

In this study, neurons were recorded and combined across the parahippocampal area while rats performed a memory-guided spatial navigation task. Sophisticated analytical tools were used to provide convincing evidence that neuronal populations in these areas show behavior-related changes that might indicate the encoding of errors by the system. The valuable results suggest that rate remapping is a likely mechanism to support changes in representations that support memory-guided behavior in these regions, most interestingly in neurons that code head direction.

Reviewer #1 (Public Review):

In this study, single neurons were recorded, using tetrodes, from the parahippocampal cortex of 5 rats navigating a double-Y maze (in which each arm of a Y-maze forks again). The goal was located at any one of the 4 branch terminations, and rats were given partial information in the form of a light cue that indicated whether the reward was on the right or left side of the maze. The second decision point was uncued and the rat had no way of knowing which of the two branches was correct, so this phase of the task was more akin to foraging. Following the outbound journey, with or without reward, the rat had to return (inbound journey) to the maze and start to begin again.

Neuronal activity was assessed for correlations with multiple navigation-relevant variables including location, head direction, speed, reward side, and goal location. The main finding is that a high proportion of neurons showed an increase in firing rate when the animal made a wrong turn at the first branch point (the one in which the correct decision was signalled). This increase, which the authors call rate remapping, persisted throughout the inbound journey as well. It was also found that head direction neurons (assessed by recording in an open field arena) in the same location in the room were more likely to show the rate change. The overall conclusion is that "during goal-directed navigation, parahippocampal neurons encode error information reflective of an animal's behavioral performance" or are "nodes in the transmission of behaviorally relevant variables during goal-directed navigation."

Overall I think this is a well-conducted study investigating an important class of neural representation: namely, the substrate for spatial orientation and navigation. The analyses are very sophisticated - possibly a little too much so, as the basic findings are relatively straightforward and the analyses take quite a bit of work to understand. A difficulty with the study is that it was exploratory (observational) rather than hypothesis-driven. Thus, the findings reveal correlations in the data but do not allow us to infer causal relationships. That said, the observation of increased firing in a subset of neurons following an erroneous choice is potentially interesting. However, the effect seems small. What were the actual firing rate values in Hz, and what was the effect size?

I also feel we are lacking information about the underlying behavior that accompanies these firing rate effects. The authors say "one possibility is that the head-direction signal in the parahippocampal region reflects a behavioral state related to the navigational choice or the lack of commitment to a particular navigational route" which is a good thought and raises the possibility that on error trials, rats are more uncertain and turn their heads more (vicarious trial and error) and thus sample the preferred firing direction more thoroughly. Another possibility is that they run more slowly, which is associated with a higher firing rate in these cells. I think we, therefore, need a better understanding of how behavior differed between error trials in terms of running speed, directional sampling, etc. A few good, convincing raw-data plots showing a remapping neuron on an error trial and a correct trial on the same arm would also be helpful (the spike plots were too tiny to get a good sense of this: fewer, larger ones would be more helpful). It would be useful to know at what point the elevated response returned to baseline, how - was it when the next trial began, and was the drop gradual (suggesting perhaps a more neurohumoral response) or sudden.

Reviewer #2 (Public Review):

This work recorded neurons in the parahippocampal regions of the medial entorhinal cortex (MEC) and pre- and para-subiculum (PrS, PaS) during a visually guided navigation task on a 'tree maze'. They found that many of the neurons reflected in their firing the visual cue (or the associated correct behavioral choice of the animal) and also the absence of reward in inbound passes (with increased firing rate). Rate remapping explained best these firing rate changes in both conditions for those cells that exhibited place-related firing. This work used a novel task, and the increased firing rate at error trials in these regions is also novel. The limitation is that cells in these regions were analyzed together.

Reviewer #3 (Public Review):

The authors set out to explore how neurons in the rodent parahippocampal area code for environmental and behavioral variables in a complex goal-directed task. The task required animals to learn the association between a cue and a spatial response and to use this information to guide behavior flexibly on a trial-by-trial basis. The authors then used a series of sophisticated analytical techniques to examine how neurons in this area encode spatial location, task-relevant cues, and correct vs. incorrect responding. While these questions have been addressed in studies of hippocampal place cells, these questions have not been addressed in these upstream parahippocampal areas.

Strengths:

1. The study presents data from ensembles of simultaneously recorded neurons in the parahippocampal region. The authors use a sophisticated method for ensuring they are not recording from the same neurons in multiple sessions and yet still report impressive sample sizes.

2. The use of the complex behavioral task guards against stereotyped behavior as rats need to continually pay attention to the relevant cue to guide behavior. The task is also quite difficult ensuring rats do not reach a ceiling level of performance which allows the authors to examine correct and incorrect trials and how spatial representations differ between them.

3. The authors take the unusual approach of not pre-processing the data to group neurons into categories based on the type of spatial information that they represent. This guards against preconceived assumptions as to how certain populations of neurons encode information.

4. The sophisticated analytical tools used throughout the manuscript allow the authors to examine spatial representations relative to a series of models of information processing.

5. The most interesting finding is that neurons in this region respond to situations where rewards are not received by increasing their firing rates. This error or mismatch signal is most commonly associated with regions of the basal ganglia and so this finding will be of particular interest to the field.

Weaknesses:

1. The histological verification of electrode position is poor and while this is acknowledged by the authors it does limit the ability to interpret these data. Recent advances have enabled researchers to look at very specific classes of neurons within traditionally defined anatomical regions and examine their interactions with well-defined targets in other parts of the brain. The lack of specificity here means that the authors have had to group MEC, PaS, and PrS into a functional group; the parahippocampus. Their primary aim is then to examine these neurons as a functional group. Given that we know that neurons in these areas differ in significant ways, there is not a strong argument for doing this.

2. The analytical/statistical tools used are very impressive but beyond the understanding of many readers. This limits the reader's ability to understand these data in reference to the rest of the literature. There are lots of places where this applies but I will describe one specific example. As noted above the authors use a complex method to examine whether neurons are recorded on multiple consecutive occasions. This is commendable as many studies in the field do not address this issue at all and it can have a major impact as analyses of multiple samples of the same neurons are often treated as if they were independent. However, there is no illustration of the outputs of this method. It would be good to see some examples of recordings that this method classifies as clearly different across days and those which are not. Some reference to previously used methods would also help the reader understand how this new method relates to those used previously.

3. The effects reported are often subtle, especially at the level of the single neuron. Examples in the figures do not support the interpretations from the population-level analysis very convincingly.

The authors largely achieve their aims with an interesting behavioral task that rats perform well but not too well. This allows them to examine memory on a trial-by-trial basis and have sufficient numbers of error trials to examine how spatial representations support memory-guided behavior. They report ensemble recordings from the parahippocampus which allows them to make conclusions about information processing within this region. This aim is relatively weak though given that this collection of areas would not usually be grouped together and treated as a single unitary area. They largely achieve their aim of examining the mechanisms underlying how these neurons code task-relevant factors such as spatial location, cue, and presence of reward. The mismatch or error-induced rate remapping will be a particularly interesting target for future research. It is also likely that the analytical tools used in this study could be used in future studies.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

In this study, single neurons were recorded, using tetrodes, from the parahippocampal cortex of 5 rats navigating a double-Y maze (in which each arm of a Y-maze forks again). The goal was located at any one of the 4 branch terminations, and rats were given partial information in the form of a light cue that indicated whether the reward was on the right or left side of the maze. The second decision point was uncued and the rat had no way of knowing which of the two branches was correct, so this phase of the task was more akin to foraging. Following the outbound journey, with or without reward, the rat had to return (inbound journey) to the maze and start to begin again.

Neuronal activity was assessed for correlations with multiple navigation-relevant variables including location, head direction, speed, reward side, and goal location. The main finding is that a high proportion of neurons showed an increase in firing rate when the animal made a wrong turn at the first branch point (the one in which the correct decision was signalled). This increase, which the authors call rate remapping, persisted throughout the inbound journey as well. It was also found that head direction neurons (assessed by recording in an open field arena) in the same location in the room were more likely to show the rate change. The overall conclusion is that "during goal-directed navigation, parahippocampal neurons encode error information reflective of an animal's behavioral performance" or are "nodes in the transmission of behaviorally relevant variables during goal-directed navigation."

Overall I think this is a well-conducted study investigating an important class of neural representation: namely, the substrate for spatial orientation and navigation. The analyses are very sophisticated - possibly a little too much so, as the basic findings are relatively straightforward and the analyses take quite a bit of work to understand. A difficulty with the study is that it was exploratory (observational) rather than hypothesis-driven. Thus, the findings reveal correlations in the data but do not allow us to infer causal relationships.

We would like to clarify that this report consists of hypothesis-driven experiments, with post-hoc exploratory analyses. We have now made hypotheses more explicit in the text, and pointed out that follow-up analyses were to understand how these effects came to be. We thank the reviewer for pointing out that our hypotheses were not explicit in the introduction. We believe our results open the door for investigating the causal role of these regions in the propagation or generation of error signals during navigational behavior. Those types of experiments are however, outside the scope of the current work.

That said, the observation of increased firing in a subset of neurons following an erroneous choice is potentially interesting. However, the effect seems small. What were the actual firing rate values in Hz, and what was the effect size?

We thank the reviewer for the opportunity to clarify the effect size question. As there are multiple neurons in the analyses, differences in firing rate need necessarily to be normalized by a neuron's mean activity. For example, a difference of 1 spk/s is less meaningful when a neuron's base rate is 50 spk/s than when it is 10spks/s. Furthermore, our reports are for population level analyses, at which point comparing raw firing rate values and differences becomes more challenging. Nonetheless, we are including these raw metrics in two new supplemental figures (Figure 2 - figure supplement 4,5), where differences in individual neurons change can be up to 15 spks/s. Additionally, the patterns and statistical results observed in the main text are preserved, with outbound Right Cue minus Left Cue showing a left>stem>right (indicating error coding), and RW minus NRW showing negative values across all segments, indicating NRW>RW or higher activity following on inbound unrewarded trials. Statistics follow the corresponding main text results (Cue: segment LRT = 71.70; RW: segment LRT=45.80).

I also feel we are lacking information about the underlying behavior that accompanies these firing rate effects. The authors say "one possibility is that the head-direction signal in the parahippocampal region reflects a behavioral state related to the navigational choice or the lack of commitment to a particular navigational route" which is a good thought and raises the possibility that on error trials, rats are more uncertain and turn their heads more (vicarious trial and error) and thus sample the preferred firing direction more thoroughly. Another possibility is that they run more slowly, which is associated with a higher firing rate in these cells. I think we, therefore, need a better understanding of how behavior differed between error trials in terms of running speed, directional sampling, etc.

In terms of running speed, there was a small effect of mean running speed between correct and incorrect trials (across subjects LMEM: Cue correct>incorrect Z=2.3, p=0.02; RW Z=2.15, p=0.03). In most neurons, increases in speed will be accompanied by increases in firing rate. Thus, the differences in running speed cannot explain the observed results, as higher speed during correct trials would predict higher activity, which is the opposite of what we found.

A few good, convincing raw-data plots showing a remapping neuron on an error trial and a correct trial on the same arm would also be helpful (the spike plots were too tiny to get a good sense of this: fewer, larger ones would be more helpful).

Additional plots for individual units have been added, Figure 2 - figure supplement 3.

It would be useful to know at what point the elevated response returned to baseline, how - was it when the next trial began, and was the drop gradual (suggesting perhaps a more neurohumoral response) or sudden.

Due to the experimental design, this question cannot be addressed fully. Concretely, error trials incur a time-penalty in which the rats need to wait an additional 10 seconds before the next trial, while a new trial would start immediately when the animal nose-poked the home well after a correct trial. Nonetheless, the data on Reward provides insight into this question. The magnitude of the responses on left and right segments of the maze were larger than on the stem for Unrewarded (NRW) vs Rewarded (RW) trials on inbound trajectories, Fig. 4c. This suggests that while activity is still elevated post-incorrect throughout the maze, across units, this effect is smaller on the stem segment. Additionally, the analyses indicate that in the transition between outbound vs inbound trajectories (Figure 4 - figure supplement 3), activity patterns are sustained (incorrect>correct). Together, these results indicate that elevated "error-like" signal are slow in returning to baseline.

Reviewer #2 (Public Review):

This work recorded neurons in the parahippocampal regions of the medial entorhinal cortex (MEC) and pre- and para-subiculum (PrS, PaS) during a visually guided navigation task on a 'tree maze'. They found that many of the neurons reflected in their firing the visual cue (or the associated correct behavioral choice of the animal) and also the absence of reward in inbound passes (with increased firing rate). Rate remapping explained best these firing rate changes in both conditions for those cells that exhibited place-related firing. This work used a novel task, and the increased firing rate at error trials in these regions is also novel. The limitation is that cells in these regions were analyzed together.

We acknowledge this limitation on our study, and we believe there might be interesting differences between these regions. Unfortunately, the post-mortem extraction of the recording implant micro-drive used for these experiments generated too much tissue damage for exact localization of the tetrodes. Nonetheless, given that the patterns were observed in all subjects, we are confident that at least the major findings of "error-like" signaling is present across the parahippocampal regions. At the same time, the distributions of functional cell types as defined in the open field are different across the PaS, PrS and MEC, leaving the possibility of a more nuanced error coding scheme by region.

Reviewer #3 (Public Review):

The authors set out to explore how neurons in the rodent parahippocampal area code for environmental and behavioral variables in a complex goal-directed task. The task required animals to learn the association between a cue and a spatial response and to use this information to guide behavior flexibly on a trial-by-trial basis. The authors then used a series of sophisticated analytical techniques to examine how neurons in this area encode spatial location, task-relevant cues, and correct vs. incorrect responding. While these questions have been addressed in studies of hippocampal place cells, these questions have not been addressed in these upstream parahippocampal areas.

Strengths:

1. The study presents data from ensembles of simultaneously recorded neurons in the parahippocampal region. The authors use a sophisticated method for ensuring they are not recording from the same neurons in multiple sessions and yet still report impressive sample sizes.

1. The use of the complex behavioral task guards against stereotyped behavior as rats need to continually pay attention to the relevant cue to guide behavior. The task is also quite difficult ensuring rats do not reach a ceiling level of performance which allows the authors to examine correct and incorrect trials and how spatial representations differ between them.

1. The authors take the unusual approach of not pre-processing the data to group neurons into categories based on the type of spatial information that they represent. This guards against preconceived assumptions as to how certain populations of neurons encode information.

1. The sophisticated analytical tools used throughout the manuscript allow the authors to examine spatial representations relative to a series of models of information processing.

1. The most interesting finding is that neurons in this region respond to situations where rewards are not received by increasing their firing rates. This error or mismatch signal is most commonly associated with regions of the basal ganglia and so this finding will be of particular interest to the field.

Weaknesses:

1. The histological verification of electrode position is poor and while this is acknowledged by the authors it does limit the ability to interpret these data. Recent advances have enabled researchers to look at very specific classes of neurons within traditionally defined anatomical regions and examine their interactions with well-defined targets in other parts of the brain. The lack of specificity here means that the authors have had to group MEC, PaS, and PrS into a functional group; the parahippocampus. Their primary aim is then to examine these neurons as a functional group. Given that we know that neurons in these areas differ in significant ways, there is not a strong argument for doing this.

See response to Reviewer 2.

1. The analytical/statistical tools used are very impressive but beyond the understanding of many readers. This limits the reader's ability to understand these data in reference to the rest of the literature. There are lots of places where this applies but I will describe one specific example. As noted above the authors use a complex method to examine whether neurons are recorded on multiple consecutive occasions. This is commendable as many studies in the field do not address this issue at all and it can have a major impact as analyses of multiple samples of the same neurons are often treated as if they were independent. However, there is no illustration of the outputs of this method. It would be good to see some examples of recordings that this method classifies as clearly different across days and those which are not. Some reference to previously used methods would also help the reader understand how this new method relates to those used previously.

We have added an additional Supplemental Figure (Figure 7 - figure supplement 1, that showcases the matching waveform approach. In the original manuscript, Fig. 7a provided an example output of the method.

1. The effects reported are often subtle, especially at the level of the single neuron. Examples in the figures do not support the interpretations from the population-level analysis very convincingly.

Additional plots for individual units have been added, Figure 2 - figure supplement 3. However, the effects, though small by unit, are consistent across neurons and subjects.

The authors largely achieve their aims with an interesting behavioral task that rats perform well but not too well. This allows them to examine memory on a trial-by-trial basis and have sufficient numbers of error trials to examine how spatial representations support memory-guided behavior. They report ensemble recordings from the parahippocampus which allows them to make conclusions about information processing within this region. This aim is relatively weak though given that this collection of areas would not usually be grouped together and treated as a single unitary area. They largely achieve their aim of examining the mechanisms underlying how these neurons code task-relevant factors such as spatial location, cue, and presence of reward. The mismatch or error-induced rate remapping will be a particularly interesting target for future research. It is also likely that the analytical tools used in this study could be used in future studies.

Reviewer #1 (Recommendations For The Authors):

1. Typo: "attempted to addresses these challenges"

We thank the reviewer for pointing out, this has been fixed.

1. "classified using tuning curve based metrics" - what does "tuning curve" mean in this context?

We have clarified this sentence in the main text.

1. "MEC neurons encode time-elapsed" should be "MEC neurons encode time elapsed" (no hyphen)

We thank the reviewer for pointing out, this has been fixed.

1. "a phenomenon referred to as 'global remapping'." - I dislike this term because it has two meanings in the literature. On the one hand, it is used to contrast with rate remapping: that is, it refers to a change in the location of place fields. On the other hand, it refers to the remapping of the whole population of cells at once, as contrasted with partial remapping. I suggest calling them location remapping (vs. rate) and complete remapping (vs. partial)

We agree that this is an overloaded term in the field. We have added 'location remapping' in the intro as a more specific term for global remapping.

1. " tasks with no trial-to-trial predictability or experimenter-controlled cues and goals in the same environment." - ambiguously worded as it isn't clear whether the "no" refers to one or both of what follows. Also needs a hyphen after experimenter.

We thank the reviewer for pointing out, this sentence has been reworded for clarity.

1. " neurons changed their firing activity as a function of cue identity" - this is confounded by behavior and reward contingency, both linked to cue identity, so the statement is slightly misleading.

We thank the reviewer for pointing this out, however, we disagree that this wording is misleading. Neurons changed their activity as a function cue identity and reward contingencies. Why neurons change their activity in such a manner is a different, more nuanced question, that we addressed through our analyses by converging on the "error" like signal that these signals seem to carry.

1. "remapping" - I am not fully comfortable with the use of this term in this context. It derives from the original reports of change in the firing location of place cells, and the proposal that these cells form a "map" with the change in activity reflecting recruitment of a new map. With observations of rate changes in some place cells, the new term "rate remapping" was introduced, and now the authors are using "rate remapping" to mean firing rate changes in non-spatial cells. The meaning is thus losing its value. "Re-coding" might be slightly better, although we can argue about whether "code" is much better than "map"

While we agree with the reviewer that "remapping" has been coerced into a grab-all term, these are the accepted semantics in the field. Thus, we are disinclined to change the language.

1. Figure 1 - it would be useful to indicate somehow that one of the decision points was cued and once free choice with the random outcome - it took me a while to work this out. Also, the choice of colors for the cues needs explaining - my understanding is that rats are very insensitive to these wavelengths. And what does Pse mean? I didn't understand those scatterplots at all.

The section "Tree-Maze behavior and electrophysiological recordings" under Results go into the details of the task. A reference and additional context for the selection of cues is now included in the "Behavioral Training" methods section. Rats possess dichromatic vision systems. Caption of Figure 1, 2 includes what Pse means, the performance of a subject for a given session. The scatter plots relate remapping to performance.

1. Also on Figure 1 - in the examples shown, it looks like the animals always checked the two end arms in the same order. Was this position habit typical?

We have added additional context in "Behavioral Training" methods section. Well trained rats do exhibit stereotyped behaviors (eg. going to one well then the other).

1. "...we hypothesized that the cue remapping score would be related to a subject's performance in the task." I am struggling to see why this doesn't follow trivially from the observation that remapping occurred on error trials.

We thank the reviewer for pointing out that this could use further clarity. We have added that the magnitude of remapping is what should relate to performance. To further clarify, remapping does not occur on error trials, remapping as operationally defined in this work, is the difference of spatial maps as a function of Cue identity or Reward contingency. Thus, as a difference metric, remapping occurs because there is a difference in activity between correct and incorrect trials. The magnitude of that difference need not relate to how the subject performed on the task.

1. "With this approach, found that incorrect coding units were more likely to overlap between cue and RW coding units than correct." Missing "we". I didn't understand this sentence - what does "overlap" mean?

We have added a sentence to further clarify this point.

1. "We found that incorrect>correct activity levels on outbound trajectories predicted incorrect>correct activity levels on inbound trajectories" - I don't understand how this can be the case given that many of these units were head direction tuned and therefore shouldn't even have been active in both directions.

As seen in Figure 7b, we were able to match 217 units across tasks. Of those, "Cluster 0" with 98 units showed strong head-direction coding. While "Cluster 0" units showed strong remapping effects, there were a lot of other units that could have contributed to the "incorrect>correct" across (out/in)-bound segments. Further, head-direction coding is defined in the Open-field environment, and there's no constraint on what these neurons could do on the Tree Maze task.

13). " Error or mismatch signals conform a fundamental computation" - should be "perform"

Wording slightly changed, but "conform" as in "act in accordance to" is what we intend here.

1. " provides it with the required stiffness and chemical resistivity"- what does "chemical resistivity" mean? To what chemicals?

This is mostly in reference to rat waste and cleaning products (alcohol). We changed the wording to durability for simplicity.

1. Supp Fig 1 shows that behavioral performance was very distinctly different for one of the animals. Was its neural data any different? What happens to the overall effect if this animal is removed from the analysis?

Unless otherwise stated, all analyses are performed through linear mixed effects with "subject" as a random effect. Thus, the effects of individual subjects are accounted for.

1. Histology - it would be useful to have a line drawing from the atlas alongside the micrographs to enable easier anatomical understanding.

There was variability in the medial lateral location of the tetrodes across animals and in the histological images provided and thus, we felt this would not be useful information as a single line drawing will not encompass/apply to all histology photos.

1. Supp. Fig. 5/6 I didn't understand what Left, Stem, and Right mean at the top. Also, the color keys are too tiny to be noticed

An additional sentence has been added to the caption to clarify that left, stem, right refer to what segment was selected via the ranking of scores.

Reviewer #2 (Recommendations For The Authors):

Was there a particular reason why cells in these regions were analyzed together? Can some of the results be tested for cells of a particular region, especially the MEC? One major limitation of these results is that it is unclear which regions it applies to, e.g., one cannot be certain that data shows here that MEC cells have these firing properties.

Damage due to the extraction of the recording tetrode bundle was extensive and we were not able to parcelate out individual regions. We have added additional details on this in the "Histology" section of the methods.

It is unclear how many cells in each region are included in each analysis. There is supplementary fig 3 but not sure how many of these met the criteria to be included in a certain analysis and it does not differentiate regions. Also, was any of the MUA included in the analyses?

Isolated single units were included in all analyses, but we did not differentiate from what region each unit came from. Analyses that include MUA are separate from the main findings, and are included in supplemental figures as reference.

Was the error trial defined as a trial when the animal did not make the right light-guided choice or did it also include cases in which the light-related arm choice was correct, but the animal first went to the unrewarded end arm? Nomenclature in the results is not explained well - what is an unrewarded trial or unrewarded trajectory or an error trial?

We have added a new paragraph in the methods under Behavioral Training that address trial nomenclature. This methods section is now referenced twice in the initial paragraphs of the results section.

Were any grid cells included in the data, especially could any cross-matched across the open field and the maze runs?

This was indeed a question of interest to us, however, the number of grid-cells recorded was not adequate for meaningful statistical inference. We further sought to avoid tuning curve based functional classifications of units.

In general, the results section is difficult to read, and its accessibility could be improved.

We thank the reviewer for this accessibility point. We hope that the small tweaks as a product of this revision will improve the readability of the manuscript. We tried to have major takeaways for each result, but the nature of the analyses necessarily make the text somewhat dense.

Minor:

One of the Figure 3f references should be Figure 3g and later, Figure 44 should be corrected.

We thank the reviewer for noting this, it has been fixed.

Reviewer #3 (Recommendations For The Authors):

There are a number of issues that I think could be addressed to improve the manuscript:

1. The figure could make it clearer where the LED panel is. Are the authors confident the rats see the cue on each trial?

We have added a new supplemental figure to address this question (Figure 1 - figure supplement 1). The new figures show the 3D geometry of the maze and the location of the Cue panel. The rats were able to see the cue, otherwise task performance would have remained at chance levels.

1. The same maze has been used in a series of studies of hippocampal place cells by Paul Dudchenko's group. They also went on to examine how these representations are affected in a very similar cued spatial response task. These studies should be acknowledged.

We thank the reviewer for pointing out this oversight. We have added the Ainge et al. citation ( https://doi.org/10.1523/JNEUROSCI.2011-07.2007) when first introducing the maze and in the methods.

1. In a number of supplementary figures, the authors present neurons that are selective for different properties such as segment, cue, reward, and direction. However, the number of spatially and cue-selective cells and the criteria by which cells are designated as selective are not reported. The analyses of spatial remapping and response to cues are done at the population level so I'm not sure how these cells are classified or selected for the figures.

The procedure for selection is included in the figure captions. Each unit is ranked based on the Uz score by segment as originally shown in Figures 2 and 4.

1. Related to this, the example cells on the figures do not clearly represent the effects presented. For example, given the title of Figure 2, I assume that the cells in 2B significantly remap. However, they don't look like they remap - the cells in the top row show rate remapping in one segment of the maze while the cells in the bottom do not show clear rate remapping responses. I suspect that traditional rate map-based analyses using maps based on consistently sized pixels rather than large segments would show only very modest changes in correlations or rates across these different types of trials. It is important to report the findings in this way as the authors interpret their data relative to the rate-remapping studies which have used these analyses. Readers who do not have the time or expertise to examine the methods in detail will conclude that the effects reported here are the same as previous rate remapping studies which the examples suggest is not the case.

Additional plots for individual units have been added to the supplement, Figure 2 - figure supplement 3. However, the effects, though small by unit, are consistent across neurons and subjects (Figure 2 - figure supplement 5).

1. Why is there a bias on the stem in 2C? This is of similar size to the effect on the right size and so deserves discussion.

The analysis in question is the across unit level bias in cue-coding by maze segment. The left segment shows elevated Right Cue coding, while the right segment shows elevated Left Cue coding. There was one reported statistical result, the main effect of segment in the Linear Mixed Effects model. We expand this result in the following two ways:

1. Individual statistical results by segment

a. Left Segment (Uz Coef. Estimate = 0.5, CI95%=[0.26, 0.75; p<1e-4])

b. Stem Segment (Uz Coef. Estimate = 0.22, CI95%=[-0.01, 0.47]; p=0.06)

c. Right Segment (Uz Coef. Estimate = -0.27, CI95%=[-0.51, -0.03], p=0.03)

1. Reporting the joint hypothesis test of left > stem > right by unit.

a. X2=90.45, p=2.28e-20

b. The comparison of left>stem by unit:

i. coefficient estimate = 0.28, CI95%=[0.11, 0.44], p=0.0008

Although the reviewer is correct in pointing out the effect size similarity, the appropriate statistical comparisons within and across units support the stated conclusions. In terms of systematic coding bias, there is a small bias across units (60% of units) and animals (4 out 5) for the Right Cue. Although interesting, this effect is orthogonal to the comparisons of interests (within unit differences). In order to highlight this point we have added the statistics of the joint hypothesis test of left>stem>right to the main manuscript.

Version 2

Abstract

Abstract

A behavioral strategy crucial to survival is directed navigation to a goal, such as a food or home location. One potential neural substrate for supporting goal-directed navigation is the parahippocampus, which contains neurons that represent an animal’s position, orientation, and movement through the world, and that change their firing activity to encode behaviorally relevant variables such as reward. However, little prior work on the parahippocampus has considered how neurons encode variables during goal-directed navigation in environments that dynamically change. Here, we recorded single units from rat parahippocampal cortex while subjects performed a goal-directed task. The maze dynamically changed goal-locations via a visual cue on a trial-to-trial basis, requiring subjects to use cue-location associations to receive reward. We observed a mismatch-like signal, with elevated neural activity on incorrect trials, leading to rate-remapping. The strength of this remapping correlated with task performance. Recordings during open-field foraging allowed us to functionally define navigational coding for a subset of the neurons recorded in the maze. This approach revealed that head-direction coding units remapped more than other functional-defined units. Taken together, this work thus raises the possibility that during goal-directed navigation, parahippocampal neurons encode error information reflective of an animal’s behavioral performance.

Introduction

Navigation to a goal location, such as a food source or home, is a behavior central to the survival of many species. This behavior begins with a serial transformation of sensory inputs and ends in a sequence of motor actions reflective of a movement decision. Between sensation and motor action, several parahippocampal regions represent variables theorized to support the computations needed for goal-directed navigation. These regions include the medial entorhinal cortex (MEC) and pre- and para-subiculum (PrS, PaS), which contain functionally defined neurons that encode an animal’s spatial position, orientation and movement through the external environment [1–6]. Additionally, the PrS and PaS project to the MEC, which is reciprocally connected to the hippocampus [7], a region strongly implicated in supporting memory and navigation [8–10]. However, much of the prior work characterizing functionally defined neurons in MEC, PrS and PaS has considered neural activity recorded as rodents forage for random food rewards in open fields environments. One limitation inherent to such open field foraging behavioral data sets is the lack of pre-defined decision points, which creates challenges to understanding the relationship between neural activity and goal-directed navigation. While several studies have attempted to address these challenges by utilizing linear environments, these works typically employed alternation tasks that lack experimental control over decision making junctures [11, 12]. Thus, how parahippocampal neurons encode navigational variables during goal-directed navigation in complex environments remains incompletely characterized.

Goal directed navigation encompasses both spatial and non-spatial coding elements, as an animal must estimate its current body position, head orientation, and running speed and integrate knowledge of behaviorally relevant goals or sensory cues. Both the MEC and PaS contain large populations of neurons that encode navigationally relevant variables for computing an animals spatial position, including: body position, head direction and running speed. Individual MEC neurons often exhibit a mixture in their selectivity for encoding these variables and, while ∼50 % of MEC neurons can be classified based on metrics that quantify the firing rate profile along a behavioral dimension (tuning curves), a larger proportion of neurons can be classified using more unbiased statistical approaches [13–15]. In the PaS, ∼50 % of neurons code for heading direction [6, 16], with coding for position or running speed present in about ∼40% neurons [15, 16]. MEC has also been shown to contain neurons that encode variables related to non-spatial features of a task or behavior. For example, subsets of MEC neurons encode time elapsed during immobility [17], change their firing rate or the spatial location at which they are maximally active in response to reward locations [18, 19], and encode an animal’s position in task-relevant space during a non-spatial auditory frequency task [20].

In addition to encoding navigationally relevant variables, parahippocampal regions have been proposed to indirectly support memory through the reciprocal projections to the hippocampal formation [9]. Depending on the task, goal-directed navigational behaviors might require knowledge of remembered cue-reward associations or memory of previous rewarded or behaviorally significant spatial locations [21–23]. At the level of neural activity, these types of memory associations may be facilitated by changes in neural activity that occur in response to changes in environmental sensory cues, such as new visual landmarks, or the presence of a reward [24, 25]. For example, in the hippocampus, place cells that are active in one or a few spatial locations will move the spatial position at which they are maximally active across different open-field environments, a phenomenon referred to as ‘global remapping’, or more specifically ‘location remapping’. Place cells can also change their firing rate, a phenomenon referred to as ‘rate remapping’, which can occur with global remapping or independent of global remapping when changes to the environment are small [24, 26–28]. In the hippocampus, both global and rate remapping have been observed in response to changes in environmental sensory cues, such as the color or shape of the spatial environment, and the presence of reward [29–31]. Remapping events observed in the hippocampus have been shown to occur in concert with changes MEC spatial firing patterns, at both the correlational level (place cell remapping correlates with re-orientation of MEC grid-cells) [32], and at the causal level (inactivation of MEC coincides with hippocampal global remapping) [33]. Both global and rate remapping measurements provide, at the population level, a quantification of stability in the neural representation of an environment through spatial correlations [24]. However, remapping might be indicative of flexibility in the representation of space, other behaviorally relevant variables, or experience [28]. Thus, quantification of remapping can be experimentally leveraged to examine whether spatial representations flexibly adapt to behaviorally relevant factors such as distinct cue-reward associations that guide decisions made during goal-directed navigation.

Neurons in MEC also exhibit global and rate remapping between different spatial environments. More-over, MEC neurons also remap in response to variables that flexibly change during goal-directed navigation, including: reward locations, running speed and an animal’s trajectory. The presence or change in the position of an unmarked remembered reward location can evoke both rate and global remapping in position and orientation encoding cells in MEC [18, 19]. Changes in an animal’s running speed have been observed to correlate with both rate remapping [34, 35] and global remapping in MEC [36]. In structured linear track mazes, in which animals perform spatial alternation tasks, MEC exhibits trajectory dependent remapping for the same spatial location [11, 12, 37]. However, while previously observed trajectory dependent remapping is indicative of prospective coding during stereotyped behavior (e.g. alternation), it remains unknown how MEC, PaS and PrS spatial patterns change in unstructured tasks (e.g. random trial-to-trial structure), or with experimenter-controlled cues and goals in the same environment. Moreover, while work in the hippocampus has shown that the degree of trajectory dependent remapping correlates with behavioral performance [38], and causal evidence revealed that shifted spatial representations correspondingly shifts the expected location of reward [21], such neural-behavior relationships have not been examined in the upstream MEC, PaS or PrS.

Here, we examined single unit neural activity in MEC, PrS, PaS as rats performed a visually cued navigational task that incorporated both goal-directed navigation and random foraging. We hypothesized that neurons in the parahippocampal formation would code for behaviorally relevant variables, like cue identity and reward. Further, we hypothesized that this neural coding could relate to the subjects performance in the task as the parahippocampal region is upstream of the hippocampus which has been shown to exhibit those relationships. These hypotheses were confirmed when we found that a large percentage of neurons changed their firing activity as a function of cue identity and reward, and that the extent of separability (as measured through remapping) in firing activity correlated with behavioral performance. To further understand how neurons changed their spatial activity patterns, we employed encoding models of rate and global remapping, where we found that changes in neural activity in the task were sufficiently explained by rate remapping. Interestingly, neural representations were also separable contingent on the presence or absence of reward at the goal well. Finally, single-unit recordings during open field foraging revealed that neurons that encoded head-direction remapped more strongly than other functionally defined units. Together, these results point to the MEC and upstream PrS/PaS as nodes in the transmission of behaviorally relevant variables during goal-directed navigation.

Results

Tree-Maze behavior and electrophysiological recordings

To examine whether MEC, PrS and PaS neurons encode variables related to goal directed navigation, we designed a linear maze with two decision junctions, referred to here as the Tree-Maze (Fig. 1, Sup. Fig. 1, see [39] and [40] for similarly structured mazes). Rats (n = 5) learned to navigate a sequence of wells to receive liquid milk rewards (Fig. 1a). There were four possible Goal-wells (G1-G4), and subjects received double rewards for triggering the correct Goal-well. Trials started at the Home-well (H). After animals triggered the Home-well (H), a visual cue turned on that indicated the correct branch to navigate after the Decision-well (D) (RC = Right-Cue, purple, Goal-wells G1-G2; LC = Left-Cue, green, Goal-wells G3-G4) (Fig. 1b). For the correct branch, only one Goal-well contained reward, which was randomly chosen on each trial. Thus, the task combined goal directed navigation (selecting the correct branch) with random foraging (selecting the first Goal-well to visit). After navigating correctly, or incorrectly (no reward and a 10 second timeout), the subject had to navigate back to the Home-well to initiate a new trial. The presence of two possible goal locations after the decisions kept subjects engaged, and dissociated the decision from the spatial location of the reward. For analyses, we segmented the maze into zones (Fig.1a), such that the location of the subject at any given time-point could be localized to a single zone (Fig. 1c).

Goal directed navigation on the Tree-Maze task a. Top view of the Tree-Maze layout and segmentation (indicated by upper and lower case letters) used for analyses (height = 1.4 m, width = 1.2 m). The LED cue panel was located at the end of the maze (flattened for illustration). Colored cue cards on the side of the maze indicate the two cue types by trial (Purple = Right-Cue [RC], Green = Left-Cue [LC]). Possible reward locations denoted by capital letters (Home = H, Decision = D, Goal = G1-G4). Lower case letters correspond to the segments shown on the right side of panel c. Reward wells highlighted in blue. b. Colored lines indicate example trajectories of a rat, 5 trials per panel. L Dec indicates a trajectory towards the left branch and R Dec indicates a trajectory towards the right branch. Left column (Left/Green cue), trajectories to G3 (top) and G4 (bottom) for reward. Both of these were correct navigational decisions and were rewarded. Right column (Right/Purple cue), trajectories to G2 (top) and G3 (bottom), only the top trajectories resulted in reward at a goal. c. Binary trajectory segmentation time window. At any given time-point, the subject can only be at one location in the maze (indicated by the white bins). Bottom axis indicates trial start times (seconds). Top axis indicates the trial number (tr), colored coded by cue. Left/Right axis indicates the identity of the segment, lower case letters on the right correspond to those in panel a. Blue and red dashed vertical lines indicate the end of a trial (blue = correct; red = incorrect). d. Task performance by subject. Each dot corresponds to a session by subject. For , dots indicates the subject mean. pse = performance in a session.

We trained 5 rats on the Tree-Maze task (see Methods-Behavioral Training). Subjects took an average of 25 sessions to reach the performance criterion for surgery (# sessions to criterion, range = 11-50; criterion: accuracy ≥70% and # trials ≥80; Sup. Fig. 2). Post-electrode implantation (Histology - Sup. Fig. 3, # units by subject Sup. Fig. 4), criteria for inclusion in the analyses were: # units ≥1, # trials ≥50 (across subject median accuracy = 72.6%, median number of trials per session = 110) (Fig. 1d). The cue on each trial was random, which allowed us to keep track of performance on Switch trials (in which the cue on trial N+1 was different than the cue on trial N, Sup. Fig. 2). Indicating that subjects were adept at the task and not simply following an alternation or exploitative strategy, the mean Switch trial performance was similar to non-Switch trials (Switch trial median accuracy = 76.9%; median number of Switch trials per session = 54). Together, the behavioral requirements of the task and structure of the Tree-Maze allowed us to then examine whether neural activity corresponded to different task dimensions (e.g. cue, decision, reward, spatial position).

Visual cues evoked spatial remapping

To investigate the effect of the cue on the spatial firing profiles of single units (i.e. neurons), we generated spatial maps by cue condition for outbound trajectories (H-well to G-well, Fig. 2a. For each unit, the spatial locations of spikes along trajectories were used to obtain trial summary metrics for each maze segment as a function of cue (Fig. 2b; Sup. Fig. 5-7). Across trials, the difference on cue coding by maze segment was quantified by the Mann-Whitney Z transformed U statistic (UZ), (Fig. 2b,c; Sup. Fig. 8, 9). In the scenario in which the cue had no effect on spatial coding, the distributions of UZ scores will be close to zero for all maze segments (Fig. 2c). On the other hand, the statistic will be negative if firing rates for the Left-Cue (LC) trials were greater than for Right-Cue (RC) trials, and positive if the firing rates for the Right-Cue (RC) trials were greater than for the Left-Cue (LC) trials. Using a linear mixed effects model we found a main effect of Segment with follow-up pairwise comparisons showing a left>stem>right pattern (LMEM; accounted for within subject variance and repeated measures for each session, Segment Likelihood Ratio Test , p = 3.07e−19; joint hypothesis test of left>stem>right Wald Test , p = 2.27e−20). Hence, across the population there was a significant bias for Right-Cue trials to be associated with greater firing rates on the left segment of the maze (relative to Left-Cue trials), and for Left-Cue trials to be associated with greater firing rates on the right segment of the maze (relative to Right-Cue trials). In both of these instances, the increased firing rate thus follows an incorrect decision, that is, trajectories that did not lead to a reward were the trajectories associated with an increased firing rate.

Spatial remapping was associated with the visual cue and correlated with task performance. a. Example session trajectories for all LC (Left-Cue) and RC (Right-Cue) trials. b. Four example single-units. For each example unit: Top row, left and middle, outbound trajectories separated by cue (red dots = spikes). Top-right, trial median activity by cue and maze segment (left, stem, right), numbers at the top indicate the Mann-Whitney Z transformed U statistic for the difference between the RC and LC trial distributions Uz. Bottom row, mean spatial rate maps by zone and cue, color coded for minimum (firing rate [FR] = 0, blue) and maximum (yellow) values. Bottom right, re-sampled distribution of correlations between RC and LC maps in blue for that unit, and in grey the corresponding null distribution. Remapping score is the mean remapping score for each unit. c. Distributions of UZ scores for all recorded units by maze segment. Purple means higher FR for RC than LC, Green means higher FR for LC than RC. Note the higher FR for RC on the left segment (far left) and higher FR for the LC on the right segment (far right). d. Distribution of mean remapping scores by unit, note the negative shift in the distribution of scores. e. Scatter-plot between the task performance on a given session pse and the cue remapping scores for recorded units in that session. Size and color of dots scale with the x axis for illustration. Regression line in red with a CI95% band. Kendall correlation score between behavior and remapping score shown. f. Scatter-plot between pse and the mean remapping score across units recorded in a given session . Size of dots indicate number of co-recorded units, color codes correspond to different subjects. Regression line and corresponding CI95% band shown in grey. g. Like (f) but with neural population correlation, composed of the spatial rate maps for all recorded units in a session. g. Correlation between pse and by subject, with bootstrapped standard deviation. is the across subject mean.

Next, to quantify the degree of spatial firing rate changes, we generated spatial zone rate maps as a function of cue. Importantly, these maps were generated from balanced re-sampling of correct and incorrect trials (see Methods -Behavioral Training). Concretely, the LC maps were generated by sampling an equal number of correct and incorrect trials from all the trials that had a LC (Fig. 2b). A distribution of scores was computed by taking the Kendall correlation, τ, between LC and RC spatial maps for each trial re-sampling instance (Fig. 2b). An appropriate null distribution was generated by sampling equal numbers of LC and RC trials from even and odd trial sets, computing the spatial correlation, and then comparing the resulting null correlation score to the cue correlation scores to generate a remapping score . Across units we observed a significant negative shift in (LMEM: mean=-0.67, p = 0.003, CI95%=(-1.105,-0.225)) (Fig. 2d). This negative shift indicated that the cue induced a significant level of remapping in firing patterns when compared to a null distribution.

Remapping of spatial representations relates to greater task performance

Given that trajectories after incorrect decisions tended to elicit an elevated FR (Fig. 2c), we hypothesized that the magnitude of the cue remapping score would related to a subject’s performance in the task. To examine this, we related the cue remapping score of each unit to the corresponding task performance for the recorded session pse (Fig. 2e). There was a significant negative relationship between performance accuracy on a given session pse and the degree of cue remapping calculated for single units (Kendall’s τ =-0.27; , CI95%=(-0.15,-0.09), , p = 8.84e 13). Given that this effect was observed for the majority of units, we hypothesized that the effect should be reflected in the mean of the cue remapping scores for co-recorded units in a session (Fig. 2f). Indeed, when considering co-recorded units in a session, the correlation between cue remapping and behavioral performance was even more pronounced (Kendall’s τ =-0.41; slope=-0.22, , p = 1.4e − 4, CI95%=(-0.33,-0.11)). Population level correlations showed the same effect, (Fig. 2g). At the individual subject level, this correlation between performance accuracy and cue remapping was observed in all but one subject, which had a small number of units (Fig. 2h). These same effects were observed on multiple versions of these analyses (Sup. Fig. 10, 11). Together, these results demonstrate cue-based remapping of parahippocampal spatial firing patterns in a goal directed navigation task, with the extent of remapping significantly correlating with behavioral accuracy in the task.

Next, we sought to understand the extent and the manner in which the visual cue was encoded in the FR of individual units during the task. We compared three different linear encoding models: (1) a zone/position model Z, (2) a cue rate-remapping model Z + C, and (3) a cue global-remapping model Z×C. The zone only model Z, simply predicted a neuron’s FR based on the zone/position the subject was occupying at a given time. The cue rate-remapping model Z + C, also received a cue identity input, acting as gain on the model to uniformly modify the predicted FR across all zones as a function of cue (Fig. 3a,b). The cue global-remapping model Z ×C doubled the number of input zones, such that only one set of zones could be active on a given cue trial. This approach thus modeled the possibility of a complete re-shuffle of the spatial firing patterns as a function of cue (Sup. Fig. 12). For each unit, the cross-validated coefficient of determination R2 was used to asses model performance (individual scores are shown in Fig. 3c). Note that a negative R2 indicates that the mean firing rate of the neuron explained more activity variance than the fitted model (on held-out test data). A subtle but consistent difference in model scores was apparent between the rate- and global remapping model (Fig. 3c,d), such that the rate remapping model outperformed the global remapping and position only model (, p = 1.24e−209) (Fig. 3d). The overall single unit scores showed a mean R2 of 12% for the Z + C model.

Cue modeling revealed rate remapping and trial-wise correlations to behavior. a. Linear zone encoding model with cue Z + C, at a given sample time t, the current position of the animal and the cue identity is multiplied by learned weights to predict FR for each recorded unit . b. Example time window of Z[t], the true FR in black V [t] and the predicted FR in red . c. Model comparison between three types of zone encoding: Z ⟶ only zones, Z + C ⟶ zones + cue, Z × C ⟶ a set of zones for each cue. Each dot is a unit, blue dots were negative R2, red scales with R2 and y-axis. d. Model comparison scores. Y-axis is the Mann-Whitney Z transformed statistic for comparing the R2 on test folds. Colorbar indicates the median difference in R2 across test folds. e. Performance of linear zone decoder models V (circles) and V +C (squares). Y-axis is the error distance in cm between the predicted and true zone, X-axis is the linearized Tree-Maze zones displayed as H (Home-well) to D (Decision-well) to i (second intersection/branching) to G (Goal-well). Linearization achieved through averaging the equivalent trajectories towards the goal. The hue shade provides groupings of sessions according to number of co-recorded units (both isolated an MUA included in these analyses). f. Performance of linear decision decoder by zone, with color indicating number of units. Note the sharp decision well split in the performance. BAC=balanced accuracy. g. Correlation between the subject’s performance and the model by zone. Color groupings as in (f.). Model performance is the comparison between the output of the decision decoder and the true identity of the cue, the same computation used to assess a subject’s performance.

To further understand how the encoding of the cue across units could modulate the strong position coding present in parahippocampal neurons, we trained a linear decoder to predict the position of the subject using the neural population activity patterns with and without the identity of the cue (V + C and V models, respectively, Fig. 3e). On held-out test data, we observed significant modulation with the cue as an input, reflected by decreased errors (the distance between the predicted and true position), but only in sessions with high number of units (LMEM: interaction between model and units groupings: , p = 1.9e−4). Additionally, the addition of cue information was most beneficial in reducing errors post-decision (LMEM: triple interaction of model, unit grouping and pre-/post-decision marker: , p = 2.9e−47). Next, we trained a classifier to predict the Left/Right decision of the subject using the population activity (Fig. 3f). Prospective decoding of the decision prior to the decision point (stem) was at chance, while it steadily increased beyond the decision point across sessions, more so in sessions with more than 8 units (LMEM: interaction of unit groupings and pre-/post decision marker, , p = 3.28e−78). The decoding of decision is reflective of the strong position coding in parahippocampal units, and also indicative of the behaviorally relevant information content of these neurons. An important note is that the number of units in a session heavily influenced the ability of the decoder to predict an upcoming Left/Right decision in the stem. Concretely, 19/42 sessions had above chance decoding with 8 or fewer co-recorded units, while 23/36 sessions with 20 or more co-recorded units had above chance decision decoding. Finally, because the decoder has trial-wise predictions of an animal’s decision Left/Right, we tested the extent to which parahippocampal representations matched the subject’s behavior (Fig. 3g). We compared decoder outputs (Left/Right decisions) to the identity of the cue, thus providing a measure comparable to the performance of subjects on the task. With this approach, we relate the performance of the subject to the model’s performance with a LMEM, with unit groupings and pre-/post-decision marker as covariates, and found a triple interaction between these predictors (, p = 3.61−10). The interaction can be broken down as follows: post-decision decoder predictions related to the subject’s performance and was modulated by the number of co-recorded neurons. These analyses demonstrate that the parahippocampal representations are not only reflective of behavior on the mean sense, as assessed through spatial remapping, but also at the individual trial-level.

Absence of reward leads to higher activity and spatial remapping

After the subjects navigated to Goal-wells, they received reward contingent on whether they made the correct navigation decision. Inbound trajectories from the Goal-wells to the Home-well could thus be separated by rewarded (correct trials) versus unrewarded (incorrect trials) (Fig. 44a). Based on previous data [18, 19], we posited that reward could also elicit firing rate changes in the spatial representations of parahippocampal neurons. We quantified firing rate changes for each unit as a function of reward by maze segment (Fig. 44b,c; Sup. Fig. 13 and Sup. Fig. 14 as a control). Across units, we found that unrewarded trials tended to have elevated firing rates across all maze segments (proportions: left=386/565, stem=340/565, right=412/565; LMEM: segment , p = 7.91e−12).

Given the firing rate differences observed between rewarded and unrewarded trials, we expected to see differences in position coding for the spatial maps corresponding to these trials sets. For each unit we generated a distribution of correlations between rewarded and unrewarded inbound spatial trajectories (bootstrapped trials with equal numbers of LC and RC). The null distribution was then the correlations between even and odd inbound trial trajectories (bootstraps having an equal number of rewarded and unrewarded trials). As with the cue, we quantified the difference between these distributions for each unit to obtain a remapping score , Fig. 4b,d. Across units, we observed a significant negative shift of the remapping score, which indicates that spatial maps changed based on the receipt or absence of reward (LMEM: mean=-1.03, CI95%=(-1.6,-0.46), p = 4.1e−4) (Fig. 4d).

Absence of reward leads to higher activity and spatial remapping. a. Example session RW (reward) and NRW (no-reward) trials. b. Four example units (i-iv). Top rows, inbound trials trajectories by RW/NRW (red dots = spikes). Top-right, trial median activity by RW and segment in the maze. Firing rate difference score Uz as described in the main text. Bottom rows, mean spatial rate maps by zone and RW, minimum FR=0. Bottom right, re-sampled distribution of correlations between RW and NRW maps in blue for that unit, and in grey the corresponding null distribution. Remapping score is the mean remapping score for each unit. b.i Session is the same as in panel a., other units from different sessions. c. Distributions of UZ scores for all recorded units by maze segment. Blues means higher FR for RW than NRW trials, Red higher NRN than RW. d. Distribution of mean remapping scores by unit, note the negative shift in the distribution of scores. e. Scatter-plot between a session’s task performance pse and the remapping scores for recorded units. Size and color of dots scale with the x axis for illustration. Regression line in red with a CI95% band. Kendall correlation score between behavior and remapping score shown. f. Scatter-plot between pse and the mean remapping score across a sessions units . Size of dots indicate number of co-recorded units, color codes correspond to different subjects. Regression line and corresponding CI95% band shown in grey. g. Like (f) but with neural population correlation, composed of the spatial rate maps for all recorded units in a session. h. Correlation between pse and by subject, with bootstrapped standard deviation. is the across subject mean.

Furthermore, because rewarded trials are correct trials, we hypothesized that as with the cue, remapping scores should relate to task performance. The extent of reward related remapping correlated with performance accuracy for a session at the level of individual units, across co-recorded units, and individual subjects (Fig. 4e-h, Sup. Fig. 15,16). Statistical analyses of these results were indicative of a strong relationship between reward remapping and task performance: unit-level (Kendall’s τ =-0.28; , CI95%=(-0.18,-0.11), , p = 3.06e−20); across co-recorded units (τ =-0.43; LMEM: slope=-0.28, CI95%=(-0.38,-0.19), , p = 3.28e−8). Similar to cue induced remapping, the extent to which units remapped in response to reward receipt significantly correlated with behavioral accuracy in the task.

As with the cue, we modeled the firing rate activity of individual neurons through three linear encoding models: (1) a zone/position model Zi, (2) a reward rate-remapping model Zi + R, and (3) a reward globalremapping model Zi × R. Where the i subscript indicates inbound trajectories, Fig. 5a,b. Similar to the cue models, we found that the activity of most units were better fitted by a rate-remapping model (Fig. 5c,d,e) (, p = 1.98e−146) (Fig. 5d). Thus these set of results indicate that parahippocampal neurons (as a population) exhibit increased firing rates during inbound trajectories that follow unrewarded trials.

Encoding model of reward remapping. a. Linear zone encoding model with reward Zi + R, at a given sample time t, the current position of the animal and the reward identity is multiplied by learned weights to predict FR for each recorded unit . b. Example time window of Z[t], the true FR in black V [t] and the predicted FR in red . c. Model comparison between three types of zone Encoding during inbound trajectories: Zi, Zi + R, Zi × R. Each dot is a unit, blue dots were negative R2, red scales with R2 and y-axis. d. Model comparison scores. Y-axis is the Mann-Whitney Z transformed statistic for comparing the R2 on test folds. Color-bar indicates the median difference in R2 across test folds.

Given the observations of spatial remapping for both cue and RW condition contingencies, we sought to understand the relationship between these results. First, we found that remapping scores were correlated (Sup. Fig. 17) across individual units and at the population level. Second, because correct/incorrect trial contingencies largely contributed to how cue and RW spatial maps were constructed, we partitioned the data into units that showed correct versus incorrect coding on the basis of their patterns of activity across maze segments (Methods). The correct/incorrect groupings were formed independently for cue (outbound) and RW (inbound) spatial maps, and we sought to find how the correct/incorrect groupings overlapped across cue and RW, (Sup. Fig. 18). With this approach, we found that incorrect coding units were more likely to overlap between cue and RW than correct coding units. However, remapping scores didn’t differ between these unit groupings. Third, we conditioned the activity of correct and incorrect inbound trajectories based on the activity of correct and incorrect outbound trajectories (Sup. Fig. 19). This analyses was motivated based on the structure of the trials: rewarded (correct) inbound trajectories followed correct outbound trajectories and that unrewarded (incorrect) inbound trajectories followed incorrect outbound trajectories. We found that incorrect>correct activity levels on outbound trajectories predicted incorrect>correct activity levels on inbound trajectories, suggesting that error signaling activity persisted through the inbound phase of the trial. Together, the remapping observations provide a quantification of this error or mismatch-like signal through different phases of the task and in the context of spatial representations.

Units encoding head direction during open field foraging are more likely to remap

In addition to the Tree-Maze, single units were recorded from subjects while they foraged for randomly scattered food rewards in a rectangular open arena (open-field foraging, OF, Fig.6a). Removable panels forming the floor of the arena were placed on top of the Tree-Maze, thus keeping peripheral visual cues relatively constant between the two environments, with the main difference coming from the 12 cm in gained height for the OF environment. A model-based encoder approach was used to quantify the extent of navigational coding for each unit along the behavioral measures of speed, head-direction or position (Fig.6).

Modeling navigational/spatial variables in neural coding during open-field foraging. a. Neural responses of four example units for subjects foraging an open-field (OF) arena [1.3m x 1.5m]. For each unit sub-panel (c0-c3): top-left, firing-rate map (number is the peak FR); top-right, head-direction tuning curve, color indicates FR magnitude by angle; bottom-left, speed tuning curve (s=speed); bottom-right, model-based variance explained on test data (R2) by variable (h=heading-direction, p=position, a=aggregate model). b. Model-based responses by variable for a selected test time-window for units c0-c3. Each row corresponds to a different model prediction , the true firing-rate (fr) for that unit, or at the top the color-coded time-window (t). Top-right, the data on which the model was trained is in grey and super-imposed is the test-window color-coded by time and with firing rate magnitude in increasing dot-size. Other heat-maps are the resulting firing-rate maps generated for the test trajectory. Model predicted rates are shown with colors matching a., with the background dotted line being the true (fr) a.. c. Population level (R2) for train (tr) and test (te) sets. d. Population level firing-rate map Pearson correlation rp between true m and predicted maps . Note that for both metrics, the aggregate model and the position model produced the best results. e. Relationship between coefficients on the aggregate model by unit. The color corresponds to the model’s training set R2.

The firing-rate predictions of these individual models was used to predict the firing rate of single units on test data in an aggregate model (Fig.6a,b). Across units, the aggregate model produced better results than individual models (LMEM: main effect of model , p = 1.33e−122, , p = 3.44−190) (Fig. 6c,d). Units across the Tree-Maze and OF tasks were matched through a waveform matching algorithm Fig. 7a (Sup. Fig. 20). In total 217 units were matched across 76 OF and 82 Tree-Maze sessions Fig. 7b. The set of model coefficients (speed, head-direction, position) for each matched unit were clustered into three groups (Fig. 7c-e). Notably, the obtained unit cluster identified a main dimension of separation between clusters in the form of a coding trade-off between head-direction and speed (Fig. 7d,e). That is, units that coded strongly for head-direction (cluster 0, blue, n=98) coded poorly for speed, while units that coded strongly for speed coded poorly for head-direction (cluster 2, green, n=59). For completeness, OF tuning score metrics were also computed for these clusters with results qualitatively matching the model coefficients (Fig. 7f).

Matched units across tasks reveal that head-direction coding units remap the strongest. a. Procedure for matching units across tasks (OF-open field, TM-tree maze). Top, 6 tetrode waveforms color-coded by matched units; bottom-left, fitted Gaussians to the dimensionally reduced unit waveforms for that tetrode across matched depth sessions (grey, unmatched units); bottom-right, symmetric confusion error matrix across units, threshold for matching PE <= 0.5. b. Venn diagram of matched units across tasks. c. UMAP clustering of the aggregate model coefficients for matched units. d. Head-direction (h) vs speed (s) coefficients formed a clustering subspace. e. Model coefficients of aggregate model used for finding clusters (p=position). Horizontal lines are the population mean, and error bars are the mean’s 95% CI. Statistics not performed, as the clusters were fitted from these parameters. f. Tuning metric scores (p=split-half rate-map correlation; s=speed-score; h=resultant-vector length score). Paired statistics through Mann-Whitney U tests (*=p < 0.05, **=p < 0.01, ***=p < 0.001, ****=p < 1e−4) h. Remap scores by OF cluster. g. TM model scores by OF cluster.

Next, we sought to relate how these clusters of units defined in the OF foraging task related to the results observed in the Tree-Maze. We observed that on this matched subset of neurons, significant remapping scores were still observed for both cue and RW contingencies, and that units in cluster 0 remapped more than the other clusters (LMEM: cluster x remap interaction , p = 0.032; cluster , p = 0.47), (Fig. 7g). Similarly, Tree-Maze encoder rate/global remapping scores were higher for cluster 0 (LMEM: cluster , p = 9.62e−4) (Fig. 7h). Additionally, we explored the extent to which these clusters related to other aspects of coding in the Tree-Maze like correct/incorrect coding and cue/RW coding by segment (Sup. Fig. 21, 22). We found that these functionally defined units had some overlap with different aspect of the maze, suggesting a small degree of heterogeneity in coding. However, most units in the clusters did not overlap with cue/RW or correct/incorrect coding. Taken together, units in cluster 0, that corresponded to stronger head-direction coding, were more prone to remap for both cue and RW than units in cluster 1 or 2. Hence, we leveraged Open-Field recordings to identify head-direction coding units as showing changes of firing patterns in the Tree-Maze, with those changes being reflective of task performance.

Discussion

During Open-Field foraging, individual MEC, PrS and PaS neurons encode an animal’s position, orientation and speed but the firing patterns of neurons in these parahippocampal regions during goal-directed navigation has been less well characterized. Here, we examined the firing patterns of hundreds of parahippocampal neurons as rats navigated a visually-guided goal directed Tree-Maze task. Rats successfully learned to use a visual cue to navigate the left or right branch of a multi-decision point linear maze to receive reward. We found that many neurons exhibited higher firing rates after incorrect left or right branch choices, which led to remapping of spatial firing rate maps. Further, the degree of remapping correlated with behavior, such that higher remapping scores corresponded to better task performance. Using encoder models, we then demonstrated that rate remapping, over global remapping, was sufficient to explain the observed changes in spatial firing rate maps. Finally, by using the navigational coefficients of encoder models fit to Open-Field foraging data, we further revealed that head direction coding cells showed the largest magnitude of remapping on the Tree-Maze. Conjunctively, these results demonstrate that parahippocampal representations of space and navigation, under goal-directed experimental conditions, project a flexible code reflective of behavior.

Neuronal encoder models serve as an important tool for identifying single-unit activity coding during open-field foraging that avoid imposing a prior on the shape of tuning curves or the generation of coding indices (e.g. tuning curve scores) [13]. Moreover, these models allow for the quantification of mixed selectivity in the coding for spatial/navigational variables, providing a more nuanced understanding of the possible variables a neural population encodes. Here, we fit linear encoder models that used speed, heading-direction and position during open-field foraging to predict each unit’s firing rate activity. Consistent with prior work, we find strong position coding, that is often complementary with coding for head-direction or speed ([1, 13, 14, 41]). Although there are differences in the proportion of cells that encode position, head direction and speed between MEC, PrS and PaS (e.g. PrS and PaS have higher proportions of head-direction tuned cells), these regions share a common substrate of spatial/navigational coding (position, head-direction, speed) [6, 15, 16]. We further found a relationship between navigation variables after clustering that revealed an axis of separation along the head-direction and speed coefficients, such that these variables were negatively related. This anti-correlation between speed and head-direction coding replicated recent observations across parahippocampal regions (MEC, PaS and PrS) [15]. Thus, as a function of speed, the head-direction signal monotonically decreases at faster speeds, consistent with the observation that subjects are less likely to turn when they run fast. Thus, one possibility is that the head-direction signal in the parahippocampal region reflects a behavioral state related to navigational choice or the lack of commitment to a particular navigational route.

Prior works point to the MEC as a key node supporting spatial decisions, as it’s inactivation led to performance deficits in visual-scene based spatial decisions [42], and subsets of MEC neurons encode the near future position of the animal (<1 second in the future) [2, 43]. Nonetheless, in terms of navigation behavior, parahippocampal neural activity after incorrect decisions has been understudied. Depending on the task, there are often very few error trials or multiple confounds that prevent substantial quantification of errors (e.g. loss of attention or satiation). Error or mismatch signals conform to a fundamental computation that informs immediate behavior, learning and neural plasticity [44, 45]. Our task structure enabled us to more rigorously consider error trials and the activity of MEC, PaS and PrS neurons before and after such trials. Given that many units exhibited higher firing rates on incorrect trials, leading to rate remapping, we hypothesized that the magnitude of remapping should correlate with behavioral performance. Indeed, we observed strong correlations between neural remapping and behavioral performance across multiple levels of analyses: individual units, across units, population correlations, individual subject level analyses and a neural population decoder model that mimicked the subject’s behavior (Figs. 2-4), as well as in control analyses that included multi-unit activity or the removal of periods of reward and/or immobility (Sup. Figs. 8-11). The robustness of this observation suggests that cortical regions encode error, expanding the possible brain regions involved in error signalling beyond downstream regions, like the basal ganglia, that prior work has shown carry error signals [44, 45].

Additional support for the idea of a mismatch signal in the parahippocampus comes from prior works that observed changes in the hippocampus related to correct or incorrect decisions. For example, hippocampal place-cells showed more efficient representations of sequences during correct trials than during incorrect [46], and non-place cells show higher rates during correct versus incorrect trials [47]. Moreover, in the T-maze, the extent of discriminability for upcoming left or right trajectories in the firing rate of hippocampal cells active on the stem of the T-maze (i.e. splitter cells) correlated with behavioral performance [38]. Of note however, we did not observe a significant population of ‘splitter cells’ in the central stem of the maze, possibly because the Tree-Maze differs from previous T- or Y-Maze in that there is a reward well at the decision point [37] [11, 12]. Even so, in our task, after training, subjects were more likely to be running on the left branch of the maze when the green visual cue (LC) was present, but after an incorrect decision, they were running on the left branch of the maze when the purple cue (RC) was present. These incorrect spatial decisions led to increase firing rates on the majority of units, and persisted on the inbound trajectories (quantified by rewarded vs unrewarded, Sup. Fig. 15). Notably, units coding for correct decisions on outbound trajectories rarely overlapped with units coding for correct decisions on inbound trajectories, while neurons coding for incorrect decisions often overlapped in their coding for inbound and outbound trajectories (Sup. Fig. 14). These observations are generally consistent with the idea that a mismatch signal could be transmitted by MEC, PaS and PrS neurons via a change in firing rate. However, it remains an open question as to where this type of mismatch signal originates such that it is broadly reflected across parahippocampal regions.

A growing set of studies have started to show that spatial and navigational codes in the parahippocampal region are more flexible than previously thought. Observations of spatial remapping in the MEC have been seen after significant changes to the environment in an open-field arena [18, 48], provoking a coordinated re-organization of grid-cells and seemingly random re-organization of other spatially selective cells. Remapping was also observed to be more likely at low running speeds [36], switching between stable spatial representations. These observations would generally be considered global remapping, as the activity fields in space change position. Other studies, however, have shown changes in firing rate and small shifts in fields near rewarded locations during foraging [18, 19]. Our data, including deep MEC, points to changes in firing rate activity on the same spatial location (rate remapping) being sufficient to explain the differences in firing rate observed between conditions (both for cue and reward). The rate remapping result is again consistent with the idea of a mismatch or error signal acting as a gain on individual neurons. Our results demonstrate that through the use of a dynamic environment we increased the behavioral demands on the animal, providing a behavioral assay to contrast parahippocampal neural activity. Further, by recording from the well characterized parahippocampal navigational circuit, we were able show that there are additional dimensions of representation or computations only elucidated through the use complex behaviors.

Supporting information